---
title: "02: SALTIRE3 scRNA-seq Atlas Rough Annotation"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to identify major cell type compartments by rough clustering of the preprocessed atlas, enabling each to be subsetted and subclustered in subsequent notebooks.

## Load packages

```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(harmony)
library(dplyr)
library(scuttle)
library(BiocParallel)
library(scDblFinder)
library(digest)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/02_S3_scRNAseq_atlas_rough_annotation/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

## Read in data

```{r}
atlas <- readRDS(file = "../03_output/01_S3_scRNAseq_preprocessing/SALTIRE3cellatlas_preprocessed.rds")
```

# Doublet detection

```{r}
# Mark known multiplets (identified by Cell Ranger Multiplexing) before running
# scDblFinder so they can be used as training doublets where available
atlas$known_doublet <- FALSE
atlas$known_doublet[which(atlas$sample_id == "Multiplet")] <- TRUE

# Stable seed per pool_id
make_seed <- function(x) {
  strtoi(substr(digest(x, algo = "xxhash32"), 1, 7), base = 16)
}

for (i in sort(unique(atlas$pool_id))) {
  
  message("Running scDblFinder for: ", i)
  
  filt <- subset(atlas, pool_id == i)
  
  filt_sce <- SingleCellExperiment(list(counts = LayerData(filt, "counts")), colData = filt@meta.data)
  
  seed_i <- make_seed(i)
  set.seed(seed_i)
  
  bp <- MulticoreParam(
    workers = 3,
    RNGseed = seed_i
  )
  
  if(i %in% c("CureAS_H5", "CureAS_U7")){
    # CureAS_H5 and CureAS_U7 are OCM pools with no hashing-based known doublets;
    # run scDblFinder without the knownDoublets argument for these
    filt_sce <- scDblFinder(
      filt_sce,
      BPPARAM = bp
    )
  } else {
    filt_sce <- scDblFinder(
      filt_sce,
      knownDoublets = "known_doublet",
      BPPARAM = bp
    )
  }
  
  scdbl_cols <- grep("^scDblFinder\\.", colnames(colData(filt_sce)), value = TRUE)
  
  atlas@meta.data[colnames(filt_sce), scdbl_cols] <-
    as.data.frame(colData(filt_sce))[, scdbl_cols, drop = FALSE]
}
```

```{r}
# Assign donor ID for OCM pools
atlas$donor_id[atlas$pool_id %in% c("CureAS_H5", "CureAS_U7")] <- "CureAS_XRU"
atlas$assay <- "scC5pV2"
atlas$assay[atlas$pool_id %in% c("CureAS_H5", "CureAS_U7")] <- "scC5pV3OCM"
```

# Clustering

```{r}
VlnPlot(atlas, features = "nCount_RNA", pt.size = 0, sort = T, group.by = "donor_id", log = T) + NoLegend()
VlnPlot(atlas, features = "nFeature_RNA", pt.size = 0, sort = T, group.by = "donor_id", log = T) + NoLegend()
VlnPlot(atlas, features = "log10GenesPerUMI", pt.size = 0, sort = T, group.by = "donor_id", log = F) + NoLegend()
VlnPlot(atlas, features = "percent_top50", pt.size = 0, sort = T, group.by = "donor_id", log = F) + NoLegend()
```

```{r}
atlas <- harmony_cluster(atlas, dims = 1:50, normalize = TRUE)
plot_cluster_umaps(atlas)
```

```{r}
# Confirm assay mixing is resolved by Harmony
DimPlot(atlas, reduction = "umap", group.by = "assay", cols = "polychrome", raster = F) + coord_equal()
```

# Marker-guided rough annotation

```{r}
Idents(atlas) <- "RNA_snn_res.0.04"
FAM(atlas, "atlas_rough_RNA_snn_res.0.04", output_path = output_path)
```

```{r}
# Canonical lineage markers: T/NK (IL7R), B (MS4A1), plasma (MZB1), myeloid (CD163),
# VIC/mesenchymal (DCN), endothelial (VWF), adipocyte (PLIN1),
# neural (GRIK2), mast (KIT), cycling (MKI67)
FeaturePlot(atlas, features = "IL7R", raster = F) + coord_equal()
FeaturePlot(atlas, features = "MS4A1", raster = F) + coord_equal()
FeaturePlot(atlas, features = "MZB1", raster = F) + coord_equal()
FeaturePlot(atlas, features = "CD163", raster = F) + coord_equal()
FeaturePlot(atlas, features = "DCN", raster = F) + coord_equal()
FeaturePlot(atlas, features = "VWF", raster = F) + coord_equal()
FeaturePlot(atlas, features = "PLIN1", raster = F) + coord_equal()
FeaturePlot(atlas, features = "GRIK2", raster = F) + coord_equal()
FeaturePlot(atlas, features = "KIT", raster = F) + coord_equal()
FeaturePlot(atlas, features = "MKI67", raster = F) + coord_equal()
```

```{r}
atlas$annotations_level1 <- ""
atlas$annotations_level1[atlas$RNA_snn_res.0.06 %in% c("0", "3")] <- "vics"
atlas$annotations_level1[atlas$RNA_snn_res.0.06 %in% c("1")] <- "myeloid"
atlas$annotations_level1[atlas$RNA_snn_res.0.06 == "2"] <- "tnk_cells"
atlas$annotations_level1[atlas$RNA_snn_res.0.06 == "4"] <- "endothelial"
atlas$annotations_level1[atlas$RNA_snn_res.0.06 %in% c("5", "7")] <- "b_cells"
atlas$annotations_level1[atlas$RNA_snn_res.0.06 == "6"] <- "mast"
atlas$annotations_level1[atlas$RNA_snn_res.0.06 == "8"] <- "cycling"
DimPlot(atlas, reduction = "umap", group.by = "annotations_level1", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
Idents(atlas) <- "annotations_level1"
FAM(atlas, name = "atlas_rough_annotations_level1", output_path = output_path)
```

# Add Cell Ranger cell-type predictions

```{r}
cell_type_meta <- distinct(atlas@meta.data, pool_id, sample_id)
cell_type_meta <- cell_type_meta[!cell_type_meta$sample_id %in%
                                   c("Multiplet", "Unassigned", "Blank", "", NA, "CureAS_OJC_LCC01", "CureAS_OJC_LCC03", "CureAS_OJC_RCC01"),] #counts too low for cell type algorithm

sample_list <- paste0(cell_type_meta$pool_id,
                      "/outs/per_sample_outs/",
                      cell_type_meta$sample_id,
                      "/count/cell_types/cell_types.csv")



write.table(sample_list,
            file = paste0(output_path, "cellranger_celltypes_file_list.txt"),
            quote = FALSE, row.names = FALSE, col.names = FALSE)
```

```{bash}
source ~/.bashrc
rsync -av -e "ssh -p 22 -i ~/.ssh/id_coconut -o IdentitiesOnly=yes" \
  --files-from=../03_output/02_S3_scRNAseq_atlas_rough_annotation/cellranger_celltypes_file_list.txt \
  coconut:"$COCONUT_CUREAS_PATH"Results/02-scnRNA-seq_Processing/02-CellRanger/ \
  ../01_data/cellranger_celltypes_output
```

```{r}
# Join per-sample Cell Ranger cell-type predictions onto the atlas metadata
atlas$predicted_coarse_cell_type <- ""
atlas$predicted_fine_cell_type <- ""
atlas$predicted_cell_count_in_model <- ""
for(i in rownames(cell_type_meta)){
  pred_path <- paste0("../01_data/cellranger_celltypes_output/", cell_type_meta[i,"pool_id"],
                      "/outs/per_sample_outs/", cell_type_meta[i,"sample_id"],
                      "/count/cell_types/cell_types.csv")
  if(file.exists(pred_path)){
    cell_type_predictions <- read.csv(pred_path, row.names = 1)
    rownames(cell_type_predictions) <- paste0(cell_type_meta[i,"pool_id"], "_", rownames(cell_type_predictions))
    overlap_barcodes <- intersect(rownames(cell_type_predictions), colnames(atlas))
    atlas@meta.data[overlap_barcodes, c("predicted_coarse_cell_type", "predicted_fine_cell_type", "predicted_cell_count_in_model")] <- cell_type_predictions[overlap_barcodes, c("coarse_cell_type", "fine_cell_type", "cell_count_in_model")]
  }
}
DimPlot(atlas, group.by = "predicted_coarse_cell_type", cols = "polychrome") + coord_equal()
DimPlot(atlas, group.by = "predicted_fine_cell_type") + coord_equal()
table(atlas$predicted_coarse_cell_type)
table(atlas$predicted_fine_cell_type)
```

# Save

```{r}
# Save per-cell-type subsets for subclustering in subsequent notebooks
for(i in unique(atlas$annotations_level1)){
  saveRDS(object = subset(atlas, annotations_level1 == i), file = paste0(output_path, i, ".rds"))
}
saveRDS(object = atlas, file = paste0(output_path, "rough_atlas.rds"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path,
                          "../01_data/output_checksums_02.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path,
                          "../01_data/output_checksums_02.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```
