---
title: "03: SALTIRE3 scRNA-seq VIC Annotation"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to annotate the valvular interstitial cell (VIC) cluster.

## Load packages

```{r}
library(Seurat)
library(SeuratDisk)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(harmony)
library(dplyr)
library(scuttle)
library(maditr)
library(openxlsx)
library(anndataR)
library(plotly)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/03_S3_scRNAseq_vic_annotation/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

## Read in data

```{r}
vics <- readRDS(file = "../03_output/02_S3_scRNAseq_atlas_rough_annotation/vics.rds")
```

# Initial clustering and QC

```{r}
vics <- harmony_cluster(vics, dims = 1:50, use_integration_features = TRUE)
plot_cluster_umaps(vics)
```

```{r}
plot_qc_by_cluster(vics, "RNA_snn_res.0.06")
```

```{r}
# Additional stress/artefact metrics to guide IEG regression in the final round
VlnPlot(vics, features = "percent_top50", group.by = "RNA_snn_res.0.06", pt.size = 0)
VlnPlot(vics, features = "log10GenesPerUMI", group.by = "RNA_snn_res.0.06", pt.size = 0)
VlnPlot(vics, features = "percent_oxphos", group.by = "RNA_snn_res.0.06", pt.size = 0)
VlnPlot(vics, features = "percent_apop", group.by = "RNA_snn_res.0.06", pt.size = 0)
VlnPlot(vics, features = "percent_dna_repair", group.by = "RNA_snn_res.0.06", pt.size = 0)
VlnPlot(vics, features = "percent_ieg", group.by = "RNA_snn_res.0.06", pt.size = 0)
VlnPlot(vics, features = "percent_ribo", group.by = "RNA_snn_res.0.06", pt.size = 0)
```

```{r}
Idents(vics) <- "RNA_snn_res.0.06"
FAM(vics, "preqc_vics_0.06", output_path = output_path)
```

## QC filtering

RNA_snn_res.0.06 cluster 2 is a mononuclear phagocyte doublet cluster; cluster 1 is very low quality.

```{r}
dim(vics)
vics <- subset(vics, RNA_snn_res.0.06 != "2" & RNA_snn_res.0.06 != "1" & known_doublet == F)
dim(vics)
```

# Second clustering round

```{r}
vics <- harmony_cluster(vics, dims = 1:50, use_integration_features = TRUE)
plot_cluster_umaps(vics)
```

```{r}
plot_qc_by_cluster(vics, "RNA_snn_res.0.4", show_doublets = FALSE)
```

```{r}
VlnPlot(vics, features = "percent_top50", group.by = "RNA_snn_res.0.4", pt.size = 0)
VlnPlot(vics, features = "log10GenesPerUMI", group.by = "RNA_snn_res.0.4", pt.size = 0)
VlnPlot(vics, features = "percent_oxphos", group.by = "RNA_snn_res.0.4", pt.size = 0)
VlnPlot(vics, features = "percent_apop", group.by = "RNA_snn_res.0.4", pt.size = 0)
VlnPlot(vics, features = "percent_dna_repair", group.by = "RNA_snn_res.0.4", pt.size = 0)
VlnPlot(vics, features = "percent_ieg", group.by = "RNA_snn_res.0.4", pt.size = 0)
VlnPlot(vics, features = "percent_ribo", group.by = "RNA_snn_res.0.4", pt.size = 0)
```

```{r}
# Interactive histograms to inspect count/feature distributions before second filter
plot_ly(x = vics$nCount_RNA, type = "histogram")
plot_ly(x = vics$nFeature_RNA, type = "histogram")
```

## Second QC filtering

Remove a low-quality cluster and apply count/feature thresholds informed by the histograms.

```{r}
dim(vics)
vics <- subset(vics, RNA_snn_res.0.4 != "6" & nCount_RNA > 1400 & nFeature_RNA > 700)
dim(vics)
```

# Final clustering round

```{r}
# Regress out IEG expression artefacts to prevent them from driving cluster structure
vics <- harmony_cluster(vics, dims = 1:50, use_integration_features = TRUE,
                        vars.to.regress = "percent_ieg")
plot_cluster_umaps(vics)
```

```{r}
# Verify VIC subtype markers resolve clearly after IEG regression
FeaturePlot(vics, features = "FAP") + coord_equal()
FeaturePlot(vics, features = "NTM") + coord_equal()
FeaturePlot(vics, features = "GFAP") + coord_equal()
```

```{r}
Idents(vics) <- "RNA_snn_res.0.1"
FAM(vics, "vics_0.1", output_path = output_path)
Idents(vics) <- "RNA_snn_res.0.3"
FAM(vics, "vics_0.3", output_path = output_path)
```

# Subclustering to resolve more VICs

```{r}
# Subcluster the largest cluster at res 0.178 to separate a neural crest-like population
Idents(vics) <- "RNA_snn_res.0.1"
vics <- FindSubCluster(vics, cluster = "0", resolution = 0.178, graph.name = "RNA_snn")
DimPlot(vics, reduction = "umap", group.by = "sub.cluster", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
Idents(vics) <- "sub.cluster"
FAM(vics, "vics_0.1_0_sub.cluster", output_path = output_path)
```

```{r}
# Further subcluster cluster 5 to separate quiescent from neural crest-like cells
Idents(vics) <- "sub.cluster"
vics <- FindSubCluster(vics, cluster = "0_1", resolution = 0.15, graph.name = "RNA_snn")
DimPlot(vics, reduction = "umap", group.by = "sub.cluster", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
DotPlot(vics, features = c("GFAP", "SOX10"), group.by = "sub.cluster")
```

# Annotation

```{r}
vics$annotations_level2_readable <- mapvalues(vics$sub.cluster, from = c("0_0","0_1_0","0_1_1","0_1_2","0_1_3","0_2","1","2","3","4"), to = c("VIC-transitional", "VIC-quiescent", "VIC-quiescent", "VIC-neural-crest", "VIC-quiescent", "VIC-FAP-osteogenic", "VIC-contractile", "VIC-spongiosa", "VIC-FAP-osteogenic", "VIC-IFN"))
vics$annotations_level2_readable <- factor(vics$annotations_level2_readable,
  levels = names(sort(table(vics$annotations_level2_readable), decreasing = T)))
vics$annotations_level2 <- mapvalues(vics$annotations_level2_readable,
  from = names(sort(table(vics$annotations_level2_readable), decreasing = T)),
  to   = paste0("VICc", 1:length(unique(vics$annotations_level2_readable))))
vics$annotations_level2 <- factor(vics$annotations_level2, levels = sort(unique(vics$annotations_level2)))
# Level 3 mirrors level 2 for VICs (no further subdivision beyond subclustering above)
vics$annotations_level3_readable <- vics$annotations_level2_readable
vics$annotations_level3 <- vics$annotations_level2

DimPlot(vics, reduction = "umap", group.by = "annotations_level2_readable", cols = "polychrome", raster = F) + coord_equal()
DimPlot(vics, reduction = "umap", group.by = "annotations_level2", cols = "polychrome", raster = F) + coord_equal()
Idents(vics) <- "annotations_level2_readable"
FAM(vics, "vics_annotations_level2_readable", output_path = output_path)
Idents(vics) <- "annotations_level3_readable"
FAM(vics, "vics_annotations_level3_readable", output_path = output_path)
```

# Save

```{r}
saveRDS(object = vics, file = paste0(output_path, "vics.rds"))
```

```{r}
# Export as h5ad for drug2cell and other Python-based analyses
vics_export <- vics
vics_export[["RNA"]] <- as(object = vics_export[["RNA"]], Class = "Assay")
anndataR::write_h5ad(vics_export, paste0(output_path, "vics_export_drug2cell.h5ad"), mode = "w")
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path,
                          "../01_data/output_checksums_03.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path,
                          "../01_data/output_checksums_03.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```
