---
title: "04: SALTIRE3 scRNA-seq Endothelial Cell Annotation"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to annotate the endothelial cell cluster.

## Load packages

```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(harmony)
library(dplyr)
library(scuttle)
library(maditr)
library(plotly)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/04_S3_scRNAseq_endothelial_annotation/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

## Read in data

```{r}
endothelial <- readRDS(file = "../03_output/02_S3_scRNAseq_atlas_rough_annotation/endothelial.rds")
```

# Initial clustering and QC

```{r}
endothelial <- harmony_cluster(endothelial, dims = 1:30, use_integration_features = TRUE)
plot_cluster_umaps(endothelial)
```

```{r}
plot_qc_by_cluster(endothelial, "RNA_snn_res.0.4")
```

```{r}
Idents(endothelial) <- "RNA_snn_res.0.2"
FAM(endothelial, "preqc_endothelial_0.2", output_path = output_path)
Idents(endothelial) <- "RNA_snn_res.0.4"
FAM(endothelial, "preqc_endothelial_0.4", output_path = output_path)
```

## QC filtering

RNA_snn_res.0.4 clusters 3 (mesenchymal), 5 (MP), and 7 (TNK) are doublet clusters; cluster 4 is low quality.

```{r}
dim(endothelial)
endothelial <- subset(endothelial, RNA_snn_res.0.4 %in% c("3", "5", "7", "4") | known_doublet == T, invert = T)
dim(endothelial)
```

# Second clustering round

```{r}
endothelial <- harmony_cluster(endothelial, dims = 1:30, use_integration_features = TRUE)
plot_cluster_umaps(endothelial)
```

```{r}
plot_qc_by_cluster(endothelial, "RNA_snn_res.0.4", show_doublets = FALSE)
```

```{r}
# Interactive histograms to inspect count/feature distributions before second filter
plot_ly(x = endothelial$nCount_RNA, type = "histogram")
plot_ly(x = endothelial$nFeature_RNA, type = "histogram")
```

```{r}
Idents(endothelial) <- "RNA_snn_res.0.4"
FAM(endothelial, "preqc2_endothelial_0.4", output_path = output_path)
```

## Second QC filtering

RNA_snn_res.0.4 cluster 4 is low quality; cluster 7 is a doublet cluster.

```{r}
dim(endothelial)
endothelial <- subset(endothelial, RNA_snn_res.0.4 %in% c("4", "7"), invert = T)
dim(endothelial)
```

# Final clustering round

```{r}
# Regress out IEG expression artefacts in the final embedding
endothelial <- harmony_cluster(endothelial, dims = 1:30, use_integration_features = TRUE,
                               vars.to.regress = "percent_ieg")
plot_cluster_umaps(endothelial)
```

```{r}
# Extend to high resolutions to help resolve rare endothelial subtypes
endothelial <- FindClusters(endothelial, resolution = c(1, 2, 3))
DimPlot(endothelial, reduction = "umap", group.by = "RNA_snn_res.1", cols = "polychrome", raster = F) + coord_equal()
DimPlot(endothelial, reduction = "umap", group.by = "RNA_snn_res.2", cols = "polychrome", raster = F) + coord_equal()
DimPlot(endothelial, reduction = "umap", group.by = "RNA_snn_res.3", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
# Subcluster cluster 4 at res 1 to resolve the lymphatic population
Idents(endothelial) <- "RNA_snn_res.1"
endothelial <- FindSubCluster(endothelial, cluster = "4", graph.name = "RNA_snn", resolution = 2)
DimPlot(endothelial, reduction = "umap", group.by = "sub.cluster", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
Idents(endothelial) <- "RNA_snn_res.0.1"
FAM(endothelial, "endothelial_0.1", output_path = output_path)
Idents(endothelial) <- "RNA_snn_res.0.2"
FAM(endothelial, "endothelial_0.2", output_path = output_path)
Idents(endothelial) <- "RNA_snn_res.0.4"
FAM(endothelial, "endothelial_0.4", output_path = output_path)
Idents(endothelial) <- "RNA_snn_res.1"
FAM(endothelial, "endothelial_1", output_path = output_path)
```

# Annotation

```{r}
# Primary assignment from res 0.4; arterial cells identified at res 0.3 cluster 5;
# lymphatics from the subcluster (1_11 = PROX1+ lymphatic endothelium)
endothelial$annotations_level3_readable <- as.character(mapvalues(endothelial$RNA_snn_res.0.3,
  from = sort(unique(endothelial$RNA_snn_res.0.3)),
  to   = c("VEC-venous","VEC-venous", "VEC-endocardial", "VEC-activated", "VEC-endocardial",
           "VEC-capillary", "VEC-arterial")))
endothelial$annotations_level3_readable[endothelial$sub.cluster == "4_9"] <- "VEC-lymphatic"
endothelial$annotations_level3_readable <- factor(endothelial$annotations_level3_readable,
  levels = names(sort(table(endothelial$annotations_level3_readable), decreasing = T)))
endothelial$annotations_level3 <- mapvalues(endothelial$annotations_level3_readable,
  from = names(sort(table(endothelial$annotations_level3_readable), decreasing = T)),
  to   = paste0("VECc", 1:length(unique(endothelial$annotations_level3_readable))))
endothelial$annotations_level3 <- factor(endothelial$annotations_level3, levels = sort(unique(endothelial$annotations_level3)))

endothelial$annotations_level2_readable <- "VEC-endocardial"
endothelial$annotations_level2_readable[endothelial$annotations_level3_readable %in% c("VEC-venous", "VEC-lymphatic", "VEC-activated", "VEC-arterial", "VEC-capillary")] <- "VEC-vascular"
endothelial$annotations_level2 <- mapvalues(endothelial$annotations_level2_readable,
  from = names(sort(table(endothelial$annotations_level2_readable), decreasing = T)),
  to   = paste0("VECc", 1:length(unique(endothelial$annotations_level2_readable))))
endothelial$annotations_level2 <- factor(endothelial$annotations_level2, levels = sort(unique(endothelial$annotations_level2)))

DimPlot(endothelial, reduction = "umap", group.by = "annotations_level2_readable", cols = "polychrome", raster = F) + coord_equal()
DimPlot(endothelial, reduction = "umap", group.by = "annotations_level3_readable", cols = "polychrome", raster = F) + coord_equal()
Idents(endothelial) <- "annotations_level2_readable"
FAM(endothelial, "endothelial_annotations_level2_readable", output_path = output_path)
Idents(endothelial) <- "annotations_level3_readable"
FAM(endothelial, "endothelial_annotations_level3_readable", output_path = output_path)
```

# Save

```{r}
saveRDS(object = endothelial, file = paste0(output_path, "endothelial.rds"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path,
                          "../01_data/output_checksums_04.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path,
                          "../01_data/output_checksums_04.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```
