---
title: "06: SALTIRE3 scRNA-seq Mononuclear Phagocyte Annotation"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to annotate the myeloid cluster.

## Load packages

```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(harmony)
library(dplyr)
library(scuttle)
library(maditr)
library(plotly)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/06_S3_scRNAseq_myeloid_annotation/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

## Read in data

```{r}
myeloid <- readRDS(file = "../03_output/02_S3_scRNAseq_atlas_rough_annotation/myeloid.rds")
```

# Initial clustering and QC

```{r}
myeloid <- harmony_cluster(myeloid, dims = 1:30, use_integration_features = TRUE)
plot_cluster_umaps(myeloid)
```

```{r}
plot_qc_by_cluster(myeloid, "RNA_snn_res.0.2")
```

```{r}
Idents(myeloid) <- "RNA_snn_res.0.2"
FAM(myeloid, "preqc_myeloid_0.2", output_path = output_path)
```

## QC filtering

RNA_snn_res.0.2 clusters 1/7/6/9/13 (mesenchyme), 4 (TNK), 11 (endothelial), and 12 (B) are doublets; clusters 2/10 are low quality.

```{r}
# Visualise cells to be removed before subsetting
myeloid$remove <- "no"
myeloid$remove[myeloid$RNA_snn_res.0.2 %in% c("1", "2", "4", "6","7","9","10","11","12","13")] <- "yes"
DimPlot(myeloid, reduction = "umap", group.by = "remove", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
dim(myeloid)
myeloid <- subset(myeloid, RNA_snn_res.0.2 %in% c("1", "2", "4", "6","7","9","10","11","12","13") | known_doublet == T, invert = T)
dim(myeloid)
```

# Second clustering round

```{r}
myeloid <- harmony_cluster(myeloid, dims = 1:30, use_integration_features = TRUE)
plot_cluster_umaps(myeloid)
```

```{r}
plot_qc_by_cluster(myeloid, "RNA_snn_res.0.3", show_doublets = FALSE)
```

```{r}
Idents(myeloid) <- "RNA_snn_res.0.3"
FAM(myeloid, "preqc2_myeloid_0.3", output_path = output_path)
```

```{r}
# Interactive histograms to inspect count/feature distributions before second filter
plot_ly(x = myeloid$nCount_RNA, type = "histogram")
plot_ly(x = myeloid$nFeature_RNA, type = "histogram")
```

## Second QC filtering

RNA_snn_res.0.3 cluster 9 is a doublet cluster; clusters 5/6 are low quality.

```{r}
dim(myeloid)
myeloid <- subset(myeloid, RNA_snn_res.0.3 %in% c("5", "6", "9"), invert = T)
dim(myeloid)
```

# Final clustering round

```{r}
# Regress out IEG expression artefacts in the final embedding
myeloid <- harmony_cluster(myeloid, dims = 1:30, use_integration_features = TRUE,
                           vars.to.regress = "percent_ieg")
plot_cluster_umaps(myeloid)
```

```{r}
Idents(myeloid) <- "RNA_snn_res.0.1"
FAM(myeloid, "myeloid_0.1", output_path = output_path)
Idents(myeloid) <- "RNA_snn_res.0.3"
FAM(myeloid, "myeloid_0.3", output_path = output_path)
Idents(myeloid) <- "RNA_snn_res.0.4"
FAM(myeloid, "myeloid_0.4", output_path = output_path)
```

# Subclustering to resolve macrophage and DC subtypes

```{r}
FeaturePlot(myeloid, features = "CTSK", order = F) + coord_equal()
```

```{r}
# Subcluster cluster 3 to separate osteoclasts (CTSK+) from resident macrophages
myeloid <- FindSubCluster(myeloid, cluster = "5", resolution = 0.1, graph.name = "RNA_snn")
DimPlot(myeloid, reduction = "umap", group.by = "sub.cluster", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
Idents(myeloid) <- "sub.cluster"
FAM(myeloid, "myeloid_sub.cluster", output_path = output_path)
```

```{r}
# Subcluster cluster 2 further to resolve DC subtypes
myeloid <- FindSubCluster(myeloid, cluster = "3", resolution = 0.2, graph.name = "RNA_snn", subcluster.name = "sub.cluster.2")
DimPlot(myeloid, reduction = "umap", group.by = "sub.cluster.2", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
Idents(myeloid) <- "sub.cluster.2"
FAM(myeloid, "myeloid_sub.cluster.2", output_path = output_path)
```

```{r}
# Subcluster cluster 2 further to resolve DC subtypes
myeloid <- FindSubCluster(myeloid, cluster = "7", resolution = 0.2, graph.name = "RNA_snn", subcluster.name = "sub.cluster.3")
DimPlot(myeloid, reduction = "umap", group.by = "sub.cluster.3", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
Idents(myeloid) <- "sub.cluster.3"
FAM(myeloid, "myeloid_sub.cluster.3", output_path = output_path)
```

```{r}
# Use Cell Ranger cell-type predictions to cross-validate cluster identity
majority_labels <- myeloid@meta.data %>%
  dplyr::group_by(sub.cluster.2) %>%
  summarise(predicted_fine_cell_type_gen = names(sort(table(predicted_fine_cell_type), decreasing = TRUE))[1])
myeloid$predicted_fine_cell_type_gen <- majority_labels$predicted_fine_cell_type_gen[match(myeloid$sub.cluster.2, majority_labels$sub.cluster.2)]
DimPlot(myeloid, reduction = "umap", group.by = "predicted_fine_cell_type_gen", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
DotPlot(myeloid, features = c("FOLR2", "LYVE1", "C1QC", "CCL4", "FCN1", "EREG", "S100A8",
                               "IL1B", "SPP1", "TREM2", "CD9", "S100P", "CSF3R", "CLEC9A",
                               "CD1C", "CTSK", "CCR7", "CD200", "FCGR3A"), cols = "RdBu") + RotatedAxis()
```

# Annotation

```{r}
myeloid$annotations_level3_readable <- ""
myeloid$annotations_level3_readable[myeloid$sub.cluster.3 %in% c("0", "1", "3_0")] <- "Resident activated macrophage"
myeloid$annotations_level3_readable[myeloid$sub.cluster.3 == "7_1"] <- "moDC"
myeloid$annotations_level3_readable[myeloid$sub.cluster.3 == "3_1"] <- "cDC2"
myeloid$annotations_level3_readable[myeloid$sub.cluster.3 == "5_0"] <- "Scar-associated macrophage"
myeloid$annotations_level3_readable[myeloid$sub.cluster.3 == "5_1"] <- "Osteoclast"
myeloid$annotations_level3_readable[myeloid$sub.cluster.3 == "2"] <- "Resident_macrophage"
myeloid$annotations_level3_readable[myeloid$sub.cluster.3 == "4"] <- "Classical_monocyte"
myeloid$annotations_level3_readable[myeloid$sub.cluster.3 == "6"] <- "Non-classical_monocyte"
myeloid$annotations_level3_readable[myeloid$sub.cluster.3 == "7_0"] <- "IFN-stim_macrophage"
myeloid$annotations_level3_readable[myeloid$sub.cluster.3 == "8"] <- "Neutrophil"
myeloid$annotations_level3_readable[myeloid$sub.cluster.3 == "9"] <- "cDC1"

myeloid$annotations_level2_readable <- ""
myeloid$annotations_level2_readable[myeloid$sub.cluster.3 %in% c("0", "1", "3_0", "5_0", "2", "7_0", "5_1")] <- "Macrophage"
myeloid$annotations_level2_readable[myeloid$sub.cluster.3 %in% c("7_1", "3_1", "9")] <- "Dendritic_cell"
myeloid$annotations_level2_readable[myeloid$sub.cluster.3 %in% c("6", "4")] <- "Monocyte"
myeloid$annotations_level2_readable[myeloid$sub.cluster.3 == "8"] <- "Neutrophil"
```

```{r}
myeloid$annotations_level2_readable <- factor(myeloid$annotations_level2_readable,
  levels = names(sort(table(myeloid$annotations_level2_readable), decreasing = T)))
myeloid$annotations_level2 <- mapvalues(myeloid$annotations_level2_readable,
  from = names(sort(table(myeloid$annotations_level2_readable), decreasing = T)),
  to   = paste0("Mc", 1:length(unique(myeloid$annotations_level2_readable))))
myeloid$annotations_level2 <- factor(myeloid$annotations_level2, levels = sort(unique(myeloid$annotations_level2)))

myeloid$annotations_level3_readable <- factor(myeloid$annotations_level3_readable,
  levels = names(sort(table(myeloid$annotations_level3_readable), decreasing = T)))
myeloid$annotations_level3 <- mapvalues(myeloid$annotations_level3_readable,
  from = names(sort(table(myeloid$annotations_level3_readable), decreasing = T)),
  to   = paste0("Mc", sprintf("%02d", 1:length(unique(myeloid$annotations_level3_readable)))))
myeloid$annotations_level3 <- factor(myeloid$annotations_level3, levels = sort(unique(myeloid$annotations_level3)))

DimPlot(myeloid, reduction = "umap", group.by = "annotations_level2_readable", cols = "polychrome", raster = F) + coord_equal()
DimPlot(myeloid, reduction = "umap", group.by = "annotations_level3_readable", cols = "polychrome", raster = F) + coord_equal()
Idents(myeloid) <- "annotations_level2_readable"
FAM(myeloid, "myeloid_annotations_level2_readable", output_path = output_path)
Idents(myeloid) <- "annotations_level3_readable"
FAM(myeloid, "myeloid_annotations_level3_readable", output_path = output_path)
```

# Save

```{r}
saveRDS(myeloid, paste0(output_path, "myeloid.rds"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path,
                          "../01_data/output_checksums_06.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path,
                          "../01_data/output_checksums_06.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```
