---
title: "07: SALTIRE3 scRNA-seq B Cell Annotation"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to annotate the B cell cluster.

## Load packages

```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(harmony)
library(dplyr)
library(scuttle)
library(maditr)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/07_S3_scRNAseq_b_cells_annotation/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

## Read in data

```{r}
b_cells <- readRDS(file = "../03_output/02_S3_scRNAseq_atlas_rough_annotation/b_cells.rds")
```

# Initial clustering and QC

```{r}
b_cells <- harmony_cluster(b_cells, dims = 1:30, use_integration_features = TRUE)
plot_cluster_umaps(b_cells)
```

```{r}
plot_qc_by_cluster(b_cells, "RNA_snn_res.0.2")
```

```{r}
Idents(b_cells) <- "RNA_snn_res.0.2"
FAM(b_cells, "preqc_b_cells_0.2", output_path = output_path)
```

## QC filtering

RNA_snn_res.0.2 clusters 3/8 (mesenchyme), 4 (myeloid), and 5 (TNK) are doublet clusters; cluster 7 is poor quality.

```{r}
dim(b_cells)
b_cells <- subset(b_cells, RNA_snn_res.0.2 %in% c("3", "4", "5", "7", "8") | known_doublet == T, invert = T)
dim(b_cells)
```

# Re-cluster after filtering

```{r}
# Regress out IEG expression artefacts in the final embedding
b_cells <- harmony_cluster(b_cells, dims = 1:30, use_integration_features = TRUE,
                           vars.to.regress = "percent_ieg")
plot_cluster_umaps(b_cells)
```

```{r}
plot_qc_by_cluster(b_cells, "RNA_snn_res.0.2", show_doublets = T)
```

```{r}
Idents(b_cells) <- "RNA_snn_res.0.2"
FAM(b_cells, "b_cells_0.2", output_path = output_path)
Idents(b_cells) <- "RNA_snn_res.0.1"
FAM(b_cells, "b_cells_0.1", output_path = output_path)
```

```{r}
# Use Cell Ranger cell-type predictions to cross-validate cluster identity
majority_labels <- b_cells@meta.data %>%
  dplyr::group_by(RNA_snn_res.0.2) %>%
  summarise(predicted_fine_cell_type_gen = names(sort(table(predicted_fine_cell_type), decreasing = TRUE))[1])
b_cells$predicted_fine_cell_type_gen <- majority_labels$predicted_fine_cell_type_gen[match(b_cells$RNA_snn_res.0.2, majority_labels$RNA_snn_res.0.2)]
DimPlot(b_cells, reduction = "umap", group.by = "predicted_fine_cell_type_gen", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
# Immunoglobulin gene expression to distinguish B cell subtypes and plasma cells
DotPlot(b_cells, features = c(
    "IGHA2", "IGHA1", "JCHAIN", "IGHM", "IGLC2", "IGLL5", "IGLC3",
    "IGHG1", "IGHG4", "IGHG3", "IGHG2", "IGHGP", "IGLC7", "IGKC"
  ), group.by = "RNA_snn_res.0.2", cols = "RdBu") + RotatedAxis() + coord_equal()
```

# Annotation

```{r}
b_cells$annotations_level3_readable <- mapvalues(b_cells$RNA_snn_res.0.2,
  from = sort(unique(b_cells$RNA_snn_res.0.2)),
  to   = c("B-naive","B-memory", "Plasma_cell", "Plasma_cell", "pDC", "B-naive"))
b_cells$annotations_level3_readable <- factor(b_cells$annotations_level3_readable,
  levels = names(sort(table(b_cells$annotations_level3_readable), decreasing = T)))
b_cells$annotations_level3 <- mapvalues(b_cells$annotations_level3_readable,
  from = names(sort(table(b_cells$annotations_level3_readable), decreasing = T)),
  to   = paste0("Bc", 1:length(unique(b_cells$annotations_level3_readable))))
b_cells$annotations_level3 <- factor(b_cells$annotations_level3, levels = sort(unique(b_cells$annotations_level3)))

b_cells$annotations_level2_readable <- mapvalues(b_cells$RNA_snn_res.0.2,
  from = sort(unique(b_cells$RNA_snn_res.0.2)),
  to   = c("B_cell","B_cell", "Plasma_cell", "Plasma_cell", "pDC", "B_cell"))
b_cells$annotations_level2_readable <- factor(b_cells$annotations_level2_readable,
  levels = names(sort(table(b_cells$annotations_level2_readable), decreasing = T)))
b_cells$annotations_level2 <- mapvalues(b_cells$annotations_level2_readable,
  from = names(sort(table(b_cells$annotations_level2_readable), decreasing = T)),
  to   = paste0("Bc", 1:length(unique(b_cells$annotations_level2_readable))))
b_cells$annotations_level2 <- factor(b_cells$annotations_level2, levels = sort(unique(b_cells$annotations_level2)))

DimPlot(b_cells, reduction = "umap", group.by = "annotations_level2_readable", cols = "polychrome", raster = F) + coord_equal()
DimPlot(b_cells, reduction = "umap", group.by = "annotations_level3_readable", cols = "polychrome", raster = F) + coord_equal()
Idents(b_cells) <- "annotations_level2_readable"
FAM(b_cells, "b_cells_annotations_level2_readable", output_path = output_path)
Idents(b_cells) <- "annotations_level3_readable"
FAM(b_cells, "b_cells_annotations_level3_readable", output_path = output_path)
```

```{r}
# Flag any residual myeloid cells that ended up in b_cells
b_cells$annotations_level1[b_cells$RNA_snn_res.0.2 %in% c("4")] <- "myeloid"
DimPlot(b_cells, reduction = "umap", group.by = "annotations_level1", cols = "polychrome", raster = F) + coord_equal()
```

# Save

```{r}
saveRDS(b_cells, paste0(output_path, "b_cells.rds"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path,
                          "../01_data/output_checksums_07.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path,
                          "../01_data/output_checksums_07.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```
