---
title: "10: SALTIRE3 Annotation Combining"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to merge the individually annotated SALTIRE3 scRNA-seq cell type objects into a single atlas, reintegrate across donors, and compute a final set of dimension reductions.

## Load packages

```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(harmony)
library(dplyr)
library(scuttle)
library(maditr)
library(readxl)
library(Matrix)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/10_S3_scRNAseq_annotations_combining/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

# Combine annotated objects

```{r}
obj_list <- list()
obj_list[[length(obj_list)+1]] <- readRDS("../03_output/03_S3_scRNAseq_vic_annotation/vics.rds")
obj_list[[length(obj_list)+1]] <- readRDS("../03_output/04_S3_scRNAseq_endothelial_annotation/endothelial.rds")
obj_list[[length(obj_list)+1]] <- readRDS("../03_output/05_S3_scRNAseq_tnk_cells_annotation/tnk_cells.rds")
obj_list[[length(obj_list)+1]] <- readRDS("../03_output/06_S3_scRNAseq_myeloid_annotation/myeloid.rds")
obj_list[[length(obj_list)+1]] <- readRDS("../03_output/07_S3_scRNAseq_b_cells_annotation/b_cells.rds")
obj_list[[length(obj_list)+1]] <- readRDS("../03_output/08_S3_scRNAseq_mast_annotation/mast.rds")
obj_list[[length(obj_list)+1]] <- readRDS("../03_output/09_S3_scRNAseq_cycling_annotation/cycling.rds")

# Annotate T and NK cells before merging
obj_list[[3]]$annotations_level1 <- "T_cell"
obj_list[[3]]$annotations_level1[obj_list[[3]]$annotations_level2_readable == "NK"] <- "NK_cell"

merged_objects <- merge(obj_list[[1]], obj_list[-1])
rm(obj_list)
```

```{r}
VlnPlot(merged_objects, features = "nCount_RNA", log = T, group.by = "donor_id", pt.size = 0) + NoLegend()
VlnPlot(merged_objects, features = "nFeature_RNA", log = T, group.by = "donor_id", pt.size = 0) + NoLegend()
VlnPlot(merged_objects, features = "nCount_RNA", log = F, group.by = "donor_id", pt.size = 0) + NoLegend()
VlnPlot(merged_objects, features = "nFeature_RNA", log = F, group.by = "donor_id", pt.size = 0) + NoLegend()
VlnPlot(merged_objects, features = "percent_mito", log = F, group.by = "donor_id", pt.size = 0) + NoLegend()
```

# Reprocess combined atlas

```{r}
# Renormalise after merge; regress out IEG artefacts for the combined embedding
merged_objects <- harmony_cluster(merged_objects, dims = 1:50, normalize = TRUE,
                                  use_integration_features = TRUE,
                                  vars.to.regress = "percent_ieg",
                                  return.model = FALSE)
plot_cluster_umaps(merged_objects)
```

```{r}
# Preview annotations carried over from subclustering notebooks
DimPlot(merged_objects, reduction = "umap", group.by = "annotations_level1", cols = "polychrome", raster = F) + coord_equal()
DimPlot(merged_objects, reduction = "umap", group.by = "annotations_level2_readable", raster = F, label = T) + coord_equal() + NoLegend()
DimPlot(merged_objects, reduction = "umap", group.by = "annotations_level3_readable", raster = F, label = T) + coord_equal() + NoLegend()
DimPlot(merged_objects, reduction = "umap", group.by = "assay", cols = "polychrome", raster = F) + coord_equal()
```

# Resolve annotation edge cases

```{r}
# Human-readable level 1 labels for plotting
merged_objects$annotations_level1_readable <- mapvalues(
  x    = merged_objects$annotations_level1,
  from = c("vics", "endothelial", "T_cell", "NK_cell", "myeloid", "b_cells", "mast", "Proliferating"),
  to   = c("Valvular interstitial cell", "Endothelial", "T cell", "NK cell",
           "Myeloid", "B cell", "Granulocyte", "Proliferating"))
merged_objects$annotations_level1_readable[merged_objects$annotations_level3_readable == "Neutrophil"] <- "Granulocyte"
merged_objects$annotations_level1[merged_objects$annotations_level1 == "mast"] <- "granulocyte"
merged_objects$annotations_level1[merged_objects$annotations_level3_readable == "Neutrophil"] <- "granulocyte"
merged_objects$annotations_level1[merged_objects$annotations_level1_readable == "Proliferating"] <- "proliferating"
merged_objects$annotations_level1[merged_objects$annotations_level3_readable == "TNK-cycling"] <- "proliferating"
merged_objects$annotations_level1_readable[merged_objects$annotations_level3_readable == "TNK-cycling"] <- "Proliferating"
merged_objects$annotations_level2_readable[merged_objects$annotations_level2_readable == "TNK-cycling"] <- "Proliferating"
merged_objects$annotations_level3_readable[merged_objects$annotations_level3_readable == "TNK-cycling"] <- "T-cycling"
merged_objects$annotations_level2_readable[merged_objects$annotations_level3_readable == "pDC"] <- "Dendritic cell"

# Level 1
level1_renames <- c(
  "Endothelial"  = "Endothelial cell"
)

merged_objects$annotations_level1_readable <- ifelse(
  merged_objects$annotations_level1_readable %in% names(level1_renames),
  level1_renames[merged_objects$annotations_level1_readable],
  merged_objects$annotations_level1_readable
)

# Level 2
level2_renames <- c(
  "Dendritic_cell" = "Dendritic cell",
  "Plasma_cell"    = "Plasma cell",
  "B_cell"         = "B cell",
  "mast"           = "Mast cell"
)

merged_objects$annotations_level2_readable <- ifelse(
  merged_objects$annotations_level2_readable %in% names(level2_renames),
  level2_renames[merged_objects$annotations_level2_readable],
  merged_objects$annotations_level2_readable
)

# Level 3
level3_renames <- c(
  "Classical_monocyte"      = "Classical monocyte",
  "Non-classical_monocyte"  = "Non-classical monocyte",
  "Resident_macrophage"     = "Resident macrophage",
  "IFN-stim_macrophage"     = "IFN-stimulated macrophage",
  "mast"                    = "Mast cell",
  "Plasma_cell"    = "Plasma cell"
)

merged_objects$annotations_level3_readable <- ifelse(
  merged_objects$annotations_level3_readable %in% names(level3_renames),
  level3_renames[merged_objects$annotations_level3_readable],
  merged_objects$annotations_level3_readable
)

DimPlot(merged_objects, reduction = "umap", group.by = "annotations_level1_readable", cols = "polychrome", raster = F) + coord_equal()
DimPlot(merged_objects, reduction = "umap", group.by = "annotations_level1", cols = "polychrome", raster = F) + coord_equal()
DimPlot(merged_objects, reduction = "umap", group.by = "annotations_level3_readable", cols = dittoColors(), raster = F) + coord_equal()
DimPlot(merged_objects, reduction = "umap", group.by = "annotations_level2_readable", cols = dittoColors(), raster = F) + coord_equal()

```

```{r}
Idents(merged_objects) <- "annotations_level1_readable"
FAM(merged_objects, "merged_objects_annotations_level1_readable", output_path = output_path)
```

```{r}
Idents(merged_objects) <- "annotations_level2_readable"
FAM(merged_objects, "merged_objects_annotations_level2_readable", output_path = output_path)
```

# Save

```{r}
saveRDS(merged_objects, file = paste0(output_path, "merged_objects.rds"))
```

```{r}
# Export as h5ad for PAGA
merged_objects_export <- merged_objects
merged_objects_export[["RNA"]] <- as(object = merged_objects_export[["RNA"]], Class = "Assay")
anndataR::write_h5ad(merged_objects_export, paste0(output_path, "merged_objects.h5ad"), mode = "w")
```

# Re-embed T/NK and myeloid subsets

Re-embed T/NK and myeloid cells in isolation to produce high-quality UMAP layouts for publication figures.

```{r}
tnk_cells <- subset(merged_objects, annotations_level1_readable == "T cell" | annotations_level1_readable == "NK cell")
tnk_cells <- harmony_cluster(tnk_cells, dims = 1:30, use_integration_features = TRUE,
                             vars.to.regress = "percent_ieg", return.model = TRUE)
DimPlot(tnk_cells, group.by = "annotations_level2_readable", cols = dittoColors()) + coord_equal()
DimPlot(tnk_cells, group.by = "annotations_level3_readable", cols = dittoColors()) + coord_equal()
saveRDS(tnk_cells, file = paste0(output_path, "tnk_cells.rds"))
```

```{r}
Idents(tnk_cells) <- "annotations_level3_readable"
FAM(tnk_cells, "tnk_cells_annotations_level3_readable", output_path = output_path)
```

```{r}
myeloid_cells <- subset(merged_objects, annotations_level1_readable == "Myeloid")
myeloid_cells <- harmony_cluster(myeloid_cells, dims = 1:30, use_integration_features = TRUE,
                                 vars.to.regress = "percent_ieg", return.model = TRUE)
DimPlot(myeloid_cells, group.by = "annotations_level2_readable", cols = dittoColors()) + coord_equal()
DimPlot(myeloid_cells, group.by = "annotations_level3_readable", cols = dittoColors()) + coord_equal()
saveRDS(myeloid_cells, file = paste0(output_path, "myeloid_cells.rds"))
```

```{r}
Idents(myeloid_cells) <- "annotations_level3_readable"
FAM(myeloid_cells, "myeloid_cells_annotations_level3_readable", output_path = output_path)
```

```{r}
merged_objects <- readRDS(file = paste0(output_path, "merged_objects.rds"))

write.table(x = data.frame(barcode_sample = Cells(merged_objects), cell_type = merged_objects$annotations_level2_readable),file = paste0(output_path, "saltire3sc_atlas_celltypes.tsv"), quote = F, sep = '\t', row.names = F)
# Save normalised counts
writeMM(merged_objects@assays$RNA$data, file = paste0(output_path, "saltire3sc_atlas.mtx"))
# save gene and cell names
write(x = rownames(merged_objects@assays$RNA$data), file = paste0(output_path, "saltire3sc_atlas_features.tsv"))
write(x = colnames(merged_objects@assays$RNA$data), file = paste0(output_path, "saltire3sc_atlas_barcodes.tsv"))
```

```{r}
# Set up reticulate to use MAGIC mamba environment
library(reticulate)
use_condaenv("scanpy-base", required = TRUE)
```

```{python}
import scanpy as sc
import pandas as pd
from scipy.io import mmread
import anndata as ad

X = mmread("../03_output/10_S3_scRNAseq_annotations_combining/saltire3sc_atlas.mtx").T.tocsr()

genes = pd.read_csv("../03_output/10_S3_scRNAseq_annotations_combining/saltire3sc_atlas_features.tsv", header=None)[0].astype(str).values
barcodes = pd.read_csv("../03_output/10_S3_scRNAseq_annotations_combining/saltire3sc_atlas_barcodes.tsv", header=None)[0].astype(str).values

adata = ad.AnnData(X=X)
adata.var_names = genes
adata.obs_names = barcodes

adata.write_h5ad("../03_output/10_S3_scRNAseq_annotations_combining/saltire3sc_atlas_counts_mtx.h5ad")
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path,
                          "../01_data/output_checksums_10.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path,
                          "../01_data/output_checksums_10.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```
