---
title: "01: UBC snRNA-seq Preprocessing"
author: "Stefan Veizades"
license: "MIT"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to perform necessary steps to load in snRNA-seq data and begin sub clustering.

## Load packages
```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(scDblFinder)
library(harmony)
library(dplyr)
library(scuttle)
library(scCustomize)
library(readxl)
library(plotly)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/01_UBC_snRNAseq_preprocessing/"
if(!dir.exists(output_path)) dir.create(output_path)
cellbender_path <- "../01_data/cellbender_output/"
souporcell_path <- "../01_data/souporcell_output/"

# Functions
source("../../utils.R")
```

# Download and read in data

## Read in sample metadata

```{r}
library_info <- as.data.frame(read_excel("../../00_study_design/library_info.xlsx"))

# Build file list for all UBC CellBender outputs to download from NAS
sample_list <- paste0(library_info$pool_id[library_info$included == "yes" & library_info$cohort == "UBC"], "/cellbender_output_filtered.h5")
write.table(sample_list, file = paste0(output_path, "cellbender_file_list.txt"), quote = F, row.names = F, col.names = F)

# Souporcell output is only generated for pooled samples (n_pool > 1);
# single-donor captures are assigned directly without demultiplexing
souporcell_list <- c(paste0(library_info$pool_id[library_info$included == "yes" & library_info$cohort == "UBC" & library_info$n_pool > 1], "/clusters.tsv"),paste0(library_info$pool_id[library_info$included == "yes" & library_info$cohort == "UBC" & library_info$n_pool > 1], "/cluster_genotypes.vcf"))
write.table(souporcell_list, file = paste0(output_path, "souporcell_file_list.txt"), quote = F, row.names = F, col.names = F)
```

```{bash}
# Retrieve cellbender/souporcell output from NAS
source ~/.bashrc
rsync -av -e "ssh -p 22 -i ~/.ssh/id_coconut -o IdentitiesOnly=yes" \
  --files-from=../03_output/01_UBC_snRNAseq_preprocessing/cellbender_file_list.txt \
  coconut:"$COCONUT_CUREAS_PATH"Results/02-scnRNA-seq_Processing/03-CellBender/ \
  ../01_data/cellbender_output

rsync -av -e "ssh -p 22 -i ~/.ssh/id_coconut -o IdentitiesOnly=yes" \
  --files-from=../03_output/01_UBC_snRNAseq_preprocessing/souporcell_file_list.txt \
  coconut:"$COCONUT_CUREAS_PATH"Results/02-scnRNA-seq_Processing/04-souporcell/ \
  ../01_data/souporcell_output
```

```{r}
# Build full paths to all downloaded inputs
cellbender_files <- paste0(cellbender_path, 
  library_info$pool_id[library_info$included == "yes" & 
                       library_info$cohort == "UBC"], 
  "/cellbender_output_filtered.h5")

souporcell_files <- paste0(souporcell_path,
  library_info$pool_id[library_info$included == "yes" & 
                       library_info$cohort == "UBC" & 
                       library_info$n_pool > 1],
  "/clusters.tsv")

all_input_files <- c(cellbender_files, souporcell_files)
```

```{r, eval=FALSE}
# Run once to generate reference checksums for non-deterministic inputs
generate_input_checksums(all_input_files, 
                         "../01_data/input_checksums_snrnaseq.csv")
```

```{r, message=TRUE}
validate_input_checksums(all_input_files, 
                         "../01_data/input_checksums_snrnaseq.csv")
```

## Read in CellBender and souporcell output

```{r}
# Iterate through included snC5pV2 (single-nucleus 5' v2) samples
obj_list <- list()
for(i in rownames(library_info)[library_info$assay == "snC5pV2" & library_info$included == "yes" & library_info$cohort == "UBC"]){
  # Read CellBender-corrected filtered count matrix
  matrix <- Read_CellBender_h5_Mat(paste0(cellbender_path, library_info[i,"pool_id"], "/cellbender_output_filtered.h5"))
  obj_list[[length(obj_list)+1]] <- CreateSeuratObject(counts = matrix, project = library_info[i,"pool_id"])
  if(library_info[i,"n_pool"] > 1){
    # For pooled captures, add souporcell SNP demultiplexing results to metadata
    souporcell_output <- read.table(paste0(souporcell_path, library_info[i,"pool_id"], "/clusters.tsv"), header = T, sep = "\t", row.names = 1)
    obj_list[[length(obj_list)]]@meta.data[rownames(souporcell_output), colnames(souporcell_output)] <- souporcell_output
  } else{
    # For single-donor captures, populate demux columns with placeholder values
    obj_list[[length(obj_list)]]$status <- rep("singlet", length(colnames(obj_list[[length(obj_list)]])))
    obj_list[[length(obj_list)]]$assignment <- rep("not_pooled", length(colnames(obj_list[[length(obj_list)]])))
  }
  # Store bare barcode before RenameCells prepends the pool_id prefix
  obj_list[[length(obj_list)]]$cell_barcode <- colnames(obj_list[[length(obj_list)]])
  obj_list[[length(obj_list)]]$pool_id <- obj_list[[length(obj_list)]]$orig.ident
  # Prepend pool_id to cell barcodes to ensure uniqueness across pools after merging
  obj_list[[length(obj_list)]] <- RenameCells(obj_list[[length(obj_list)]], add.cell.id = library_info[i,"pool_id"])
}

# Merge all pool objects into a single atlas; JoinLayers collapses the per-sample
# count layers created by Seurat v5 into a single counts layer
atlas <- merge(obj_list[[1]], c(obj_list[2:length(obj_list)]))
atlas <- JoinLayers(atlas)
```

# QC metrics and base filtering

## Add QC metrics

```{r}
atlas <- Add_Cell_QC_Metrics(object = atlas, species = "human")
QC_Plots_Genes(seurat_object = atlas, pt.size = 0, raster = F)
QC_Plots_UMIs(seurat_object = atlas, pt.size = 0, raster = F)
QC_Plots_Mito(seurat_object = atlas, pt.size = 0, raster = F)
QC_Plots_Complexity(seurat_object = atlas, pt.size = 0, raster = F)
```

## Interactively evaluate distribution of quality metrics

```{r}
plot_ly(x = atlas$nCount_RNA, type = "histogram")
plot_ly(x = atlas$nFeature_RNA, type = "histogram")
plot_ly(x = atlas$percent_mito, type = "histogram")
plot_ly(x = atlas$percent_ribo, type = "histogram")
plot_ly(x = atlas$log10GenesPerUMI, type = "histogram")
```

## Calculate outlier value for QC metrics

```{r}
# Calculate data-adaptive QC thresholds using the median absolute deviation (MAD) method.
# Cells >3 MADs below the median UMI/gene count, or >3 MADs above the median
# mitochondrial fraction, are flagged as outliers.
attr(isOutlier(atlas$nCount_RNA, nmads = 3, type = "lower", log = T),"thresholds")
attr(isOutlier(atlas$nFeature_RNA, nmads = 3, type = "lower", log = T),"thresholds")
attr(isOutlier(atlas$percent_mito, nmads = 3, type = "higher", log = F),"thresholds")
```

## Visualise cut offs

```{r}
QC_Histogram(seurat_object = atlas, features = "nCount_RNA", low_cutoff = 360.1015)
QC_Histogram(seurat_object = atlas, features = "nFeature_RNA", low_cutoff = 380.7693)
QC_Histogram(seurat_object = atlas, features = "percent_mito", high_cutoff = 1) # manually set for very low calculated cut off
```

```{r}
QC_Plot_UMIvsGene(seurat_object = atlas, low_cutoff_gene = 384.5321, low_cutoff_UMI = 360.1015, raster = F)
QC_Plot_GenevsFeature(seurat_object = atlas, feature1 = "percent_mito", low_cutoff_gene = 380.7693, high_cutoff_feature = 1, raster = F)
```

## Remove QC outliers

```{r}
# Flag outlier cells per metric
atlas$nCount_RNA_outlier <- isOutlier(atlas$nCount_RNA, nmads = 3, type = "lower", log = T)
atlas$nFeature_RNA_outlier <- isOutlier(atlas$nFeature_RNA, nmads = 3, type = "lower", log = T)
atlas$percent.mt_outlier <- atlas$percent_mito>1  # manual threshold; MAD-derived cutoff was <1%

# Save unfiltered object before subsetting, to allow inspection of removed cells
saveRDS(atlas, file = paste0(output_path, "UBCnucatlas_unprocessed.rds"))
atlas <- subset(atlas, subset = nCount_RNA_outlier == FALSE & nFeature_RNA_outlier == FALSE & percent.mt_outlier == FALSE)
```

```{r}
VlnPlot(atlas, features = c("nFeature_RNA", "nCount_RNA", "percent_mito"), ncol = 3, log = T, pt.size = 0, raster = F)& theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+NoLegend()
VlnPlot(atlas, features = c("nFeature_RNA", "nCount_RNA", "percent_mito"), ncol = 3, log = F, pt.size = 0, raster = F)& theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+NoLegend()
QC_Plot_UMIvsGene(seurat_object = atlas, raster = F)
QC_Plot_GenevsFeature(seurat_object = atlas, feature1 = "percent_mito", raster = F)
```

```{r}
saveRDS(atlas, file = paste0(output_path, "UBCnucatlas_preprocessed.rds"))
```

# Prepare velocyto input

```{r}
# Write per-pool filtered count matrices for velocyto input.
# Velocyto requires 10x-formatted files (barcodes, features, matrix) and the
# original BAM, so we stage the filtered barcodes here for use on the HPC.
full_matrix <- atlas@assays$RNA@layers$counts
base_path <- paste0(output_path,"velocyto_input/")
if(!dir.exists(base_path)) dir.create(base_path)
for(i in unique(atlas$pool_id)){
  write10xCounts(path = paste0(base_path,i),
                 x = full_matrix[,atlas$pool_id == i],
                 # Strip the pool_id prefix (first 10 chars + "_") added by RenameCells
                 # to recover the original 10x cell barcodes expected by velocyto
                 barcodes = substring(Cells(atlas)[atlas$pool_id == i],first = 11),
                 gene.id = rownames(atlas),
                 version = "3", overwrite = T)
}
# Generate velocyto HPC submission scripts for each pool
for(i in unique(atlas$pool_id)){
  velocyto(pool_id = i, output_path = paste0(base_path))
}
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path, 
                          "../01_data/output_checksums_01.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path, 
                          "../01_data/output_checksums_01.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```
