---
title: "02: UBC snRNA-seq SNP Demultiplexing"
author: "Stefan Veizades"
license: "MIT"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to assign donor IDs to cells using SNP genotyping concordance. For each pooled capture, souporcell cluster genotypes are compared against a reference WGS VCF using pairwise genotype concordance, and the best-matching donor is assigned to each cluster. Single-donor captures are assigned directly from the pool metadata.

## Load packages

```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(scDblFinder)
library(harmony)
library(dplyr)
library(scuttle)
library(viridis)
library(RColorBrewer)
library(patchwork)
library(VariantAnnotation)
library(GenomicRanges)
library(pheatmap)
library(readxl)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/02_UBC_snRNAseq_dehashing/"
if(!dir.exists(output_path)) dir.create(output_path)
souporcell_path <- "../01_data/souporcell_output/"

# Functions
source("../../utils.R")
```

## Download and read in data

```{r}
# UBC_patient_pools maps each pool_id to the donor IDs it contains
UBC_patient_pools <- as.data.frame(read_xlsx("../../00_study_design/UBC_patient_pools.xlsx"))
atlas <- readRDS("../03_output/01_UBC_snRNAseq_preprocessing/UBCnucatlas_preprocessed.rds")
# Load reference WGS genotypes for all UBC donors
reference_vcf <- load_vcf_genotypes("../../00_study_design/genotype_data.vcf")
# Prepend "chr" to match the chromosome naming convention used by souporcell VCFs
rownames(reference_vcf) <- paste0("chr", rownames(reference_vcf))
```

# Demultiplex pooled samples

## Preset values

```{r}
# souporcell encodes cell assignment as a cluster index ("0", "1", ...), "doublet",
# or "unassigned"; not_pooled cells were set to "not_pooled" in preprocessing.
# Taking the first character gives a compact key: digit = singlet cluster,
# "d" = doublet, "u" = unassigned, "n" = not_pooled (single-donor capture).
atlas$singlet_holder <- substring(atlas$assignment, first = 1, last = 1)
atlas$multiplex_id <- paste0(atlas$pool_id, "_", atlas$singlet_holder)
table(atlas$multiplex_id)
```

## Iterate through pools and assign donor matches

```{r}
atlas$donor_id <- ""

# Process pooled captures (singlet_holder != "n", i.e. not single-donor captures).
# For each pool, match souporcell cluster genotypes to reference WGS genotypes by
# pairwise concordance, then map cluster indices to donor IDs.
# Doublet/unassigned cells pass through mapvalues unchanged (no matching key found).
for(i in unique(atlas$pool_id[atlas$singlet_holder != "n"])){
  souporcell_vcf <- load_vcf_genotypes(paste0(souporcell_path, i, "/cluster_genotypes.vcf"))
  # Restrict reference to donors known to be in this pool
  reference_vcf_subset <- reference_vcf[, colnames(reference_vcf) %in% UBC_patient_pools$donor_id[UBC_patient_pools$pool == i]]
  results <- match_donors(souporcell_vcf, reference_vcf_subset)
  print(paste0(i, " Dehashing"))
  print("Similarity Matrix:")
  print(results$SimilarityMatrix)
  print("Donor Matches:")
  print(results$DonorMatches)
  # Save similarity matrix for manual review
  heatmap <- pheatmap(
      results$SimilarityMatrix,
      cluster_rows = TRUE,
      cluster_cols = TRUE,
      main = "Heatmap of Donor Similarity",
      display_numbers = TRUE,
      fontsize_number = 10
  )
  ggsave(paste0(output_path, i, "_donor_similarity_matrix.pdf"), plot = heatmap)
  write.csv(results$SimilarityMatrix, file = paste0(output_path, i, "_donor_similarity_matrix.csv"))
  # Map souporcell cluster index → matched donor ID for all cells in this pool
  atlas$donor_id[atlas$pool_id == i] <- mapvalues(atlas$singlet_holder[atlas$pool_id == i], from = results$DonorMatches$Donor1, to = results$DonorMatches$Donor2)
}

# For single-donor captures, assign donor ID directly from pool metadata
atlas$donor_id[atlas$singlet_holder == "n"] <- mapvalues(atlas$pool_id[atlas$singlet_holder == "n"], from = UBC_patient_pools$pool[UBC_patient_pools$pool %in%  unique(atlas$pool_id[atlas$singlet_holder == "n"])], to = UBC_patient_pools$donor_id[UBC_patient_pools$pool %in%  unique(atlas$pool_id[atlas$singlet_holder == "n"])])
table(atlas$donor_id)
```

## Save demultiplexed atlas

```{r}
saveRDS(atlas, file = paste0(output_path, "dehashed_atlas.rds"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path, 
                          "../01_data/output_checksums_02.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path, 
                          "../01_data/output_checksums_02.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```