---
title: "03: UBC snRNA-seq Atlas Rough Annotation"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to perform an initial broad clustering of the full atlas to identify major cell type groups, which are then subsetted for cell-type-specific annotation in subsequent notebooks.

## Load packages

```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(harmony)
library(dplyr)
library(scuttle)
library(BiocParallel)
library(scDblFinder)
library(readxl)
library(digest)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/03_UBC_snRNAseq_atlas_rough_annotation/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

## Read in data

```{r}
atlas <- readRDS(file = "../03_output/02_UBC_snRNAseq_dehashing/dehashed_atlas.rds")
donor_info <- as.data.frame(read_excel("../../00_study_design/patient_information.xlsx"))
```

```{r}
# Map donor IDs to clinical classification
atlas$clinical_classification <- mapvalues(atlas$donor_id, from = donor_info$donor_id, to = donor_info$clinical_classification)
atlas$clinical_classification_readable <- mapvalues(atlas$donor_id, from = donor_info$donor_id, to = donor_info$clinical_classification_readable)
table(atlas$clinical_classification)
table(atlas$clinical_classification_readable)
```

## Doublet detection

```{r}
# Flag souporcell-identified doublets, then run scDblFinder per pool.
# Pools without souporcell demux (single-donor captures) are run unsupervised.
atlas$known_doublet <- atlas$status == "doublet"

# Stable seed per pool_id
make_seed <- function(x) {
  strtoi(substr(digest(x, algo = "xxhash32"), 1, 7), base = 16)
}

for (i in sort(unique(atlas$pool_id))) {
  
  message("Running scDblFinder for: ", i)
  
  filt <- subset(atlas, pool_id == i)
  
  filt_sce <- SingleCellExperiment(
    list(counts = filt@assays$RNA@layers$counts),
    colData = filt@meta.data
  )
  
  seed_i <- make_seed(i)
  set.seed(seed_i)
  
  bp <- MulticoreParam(
    workers = 3,
    RNGseed = seed_i
  )
  
  if (i %in% c("CureAS_QH", "CureAS_YL", "CureAS_KQ", "CureAS_YA")) {
    # Single-donor captures: no souporcell doublets available, run unsupervised
    filt_sce <- scDblFinder(
      filt_sce,
      BPPARAM = bp
    )
  } else {
    filt_sce <- scDblFinder(
      filt_sce,
      knownDoublets = "known_doublet",
      BPPARAM = bp
    )
  }
  
  scdbl_cols <- grep("^scDblFinder\\.", colnames(colData(filt_sce)), value = TRUE)
  
  atlas@meta.data[colnames(filt_sce), scdbl_cols] <-
    as.data.frame(colData(filt_sce))[, scdbl_cols, drop = FALSE]
}
```

```{r}
# Review per-donor QC metrics to identify low-quality donors for exclusion
VlnPlot(atlas, features = "nCount_RNA", pt.size = 0, sort = T, group.by = "donor_id", log = T) + NoLegend()
VlnPlot(atlas, features = "nFeature_RNA", pt.size = 0, sort = T, group.by = "donor_id", log = T) + NoLegend()
VlnPlot(atlas, features = "log10GenesPerUMI", pt.size = 0, sort = T, group.by = "donor_id", log = F) + NoLegend()
VlnPlot(atlas, features = "percent_top50", pt.size = 0, sort = T, group.by = "donor_id", log = F) + NoLegend()
```

```{r}
# Exclude three donors with poor quality metrics, then normalise, integrate
# across donors with Harmony, and cluster across a range of resolutions
atlas <- subset(atlas, donor_id %in% c("CureAS_KS4", "CureAS_C4J", "CureAS_PUM"), invert = T)
atlas <- harmony_cluster(atlas, dims = 1:50, normalize = TRUE)
plot_cluster_umaps(atlas)
```

```{r}
# Canonical marker genes for broad cell type identification
FeaturePlot(atlas, features = "IL7R", raster = F) + coord_equal()   # T cells
FeaturePlot(atlas, features = "MS4A1", raster = F) + coord_equal()  # B cells
FeaturePlot(atlas, features = "CD163", raster = F) + coord_equal()  # Macrophages/monocytes
FeaturePlot(atlas, features = "DCN", raster = F) + coord_equal()    # VICs
FeaturePlot(atlas, features = "VWF", raster = F) + coord_equal()    # Endothelial
FeaturePlot(atlas, features = "PLIN1", raster = F) + coord_equal()  # Adipocytes
FeaturePlot(atlas, features = "GRIK2", raster = F) + coord_equal()  # Neuronal
FeaturePlot(atlas, features = "KIT", raster = F) + coord_equal()    # Mast cells
FeaturePlot(atlas, features = "MKI67", raster = F) + coord_equal()  # Cycling cells
```

```{r}
# Assign broad cell type labels at resolution 0.02; cycling cells require
# resolution 0.06 to separate from the VIC cluster
atlas$annotations_level1 <- ""
atlas$annotations_level1[atlas$RNA_snn_res.0.02 == "0"] <- "vics"
atlas$annotations_level1[atlas$RNA_snn_res.0.02 == "1"] <- "myeloid"
atlas$annotations_level1[atlas$RNA_snn_res.0.02 == "2"] <- "tnkb_cells"
atlas$annotations_level1[atlas$RNA_snn_res.0.02 == "3"] <- "endothelial"
atlas$annotations_level1[atlas$RNA_snn_res.0.02 == "4"] <- "adipocytes"
atlas$annotations_level1[atlas$RNA_snn_res.0.02 == "5"] <- "mast"
atlas$annotations_level1[atlas$RNA_snn_res.0.06 == "7"] <- "cycling"
DimPlot(atlas, reduction = "umap", group.by = "annotations_level1", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
Idents(atlas) <- "annotations_level1"
FAM(atlas, name = "atlas_rough_annotations_level1", output_path = output_path)
```

# Load Cell Ranger cell type predictions

```{r}
sample_list <- paste0(unique(atlas$pool_id), "/outs/cell_types/cell_types.csv")
write.table(sample_list, file = paste0(output_path, "cellranger_celltypes_file_list.txt"), quote = F, row.names = F, col.names = F)
```

```{bash}
# Retrieve cellranger cell type output from NAS
source ~/.bashrc
rsync -av -e "ssh -p 22 -i ~/.ssh/id_coconut -o IdentitiesOnly=yes" \
  --files-from=../03_output/03_UBC_snRNAseq_atlas_rough_annotation/cellranger_celltypes_file_list.txt \
  coconut:"$COCONUT_CUREAS_PATH"Results/02-scnRNA-seq_Processing/02-CellRanger/ \
  ../01_data/cellranger_celltypes_output
```

```{r}
# Build full paths to all downloaded inputs
celltype_files <- paste0("../01_data/cellranger_celltypes_output/", unique(atlas$pool_id), "/outs/cell_types/cell_types.csv")
```

```{r, eval=FALSE}
# Run once to generate reference checksums for non-deterministic inputs
generate_input_checksums(celltype_files, 
                         "../01_data/input_checksums_celltypes.csv")
```

```{r, message=TRUE}
validate_input_checksums(celltype_files, 
                         "../01_data/input_checksums_celltypes.csv")
```

```{r}
# Add Cell Ranger automated cell type predictions as metadata for comparison with
# manual annotations; overlap_barcodes handles cells absent from prediction output
atlas$predicted_coarse_cell_type <- ""
atlas$predicted_fine_cell_type <- ""
atlas$predicted_cell_count_in_model <- ""
for(i in unique(atlas$pool_id)){
  cell_type_predictions <- read.csv(
    paste0("../01_data/cellranger_celltypes_output/", i, "/outs/cell_types/cell_types.csv"),
    row.names = 1
  )
  rownames(cell_type_predictions) <- paste0(i, "_", rownames(cell_type_predictions))
  overlap_barcodes <- intersect(rownames(cell_type_predictions), colnames(atlas))
  atlas@meta.data[overlap_barcodes, c("predicted_coarse_cell_type", "predicted_fine_cell_type", "predicted_cell_count_in_model")] <-
    cell_type_predictions[overlap_barcodes, c("coarse_cell_type", "fine_cell_type", "cell_count_in_model")]
}
DimPlot(atlas, group.by = "predicted_coarse_cell_type", cols = "polychrome") + coord_equal()
DimPlot(atlas, group.by = "predicted_fine_cell_type") + coord_equal()
table(atlas$predicted_coarse_cell_type)
table(atlas$predicted_fine_cell_type)
```

# Save

```{r}
# Save each broad cell type as a separate RDS for subclustering in notebooks 04-10
for(i in unique(atlas$annotations_level1)){
  saveRDS(object = subset(atlas, annotations_level1 == i), file = paste0(output_path, i, ".rds"))
}
saveRDS(object = atlas, file = paste0(output_path, "rough_atlas.rds"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path, 
                          "../01_data/output_checksums_03.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path, 
                          "../01_data/output_checksums_03.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```
