---
title: "04: UBC snRNA-seq VIC Annotation"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to annotate the valvular interstitial cell (VIC) cluster. VICs are the predominant stromal cell type of the aortic valve.

## Load packages

```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(harmony)
library(dplyr)
library(scuttle)
library(maditr)
library(org.Hs.eg.db)
library(clusterProfiler)
library(biomaRt)
library(openxlsx)
library(SeuratDisk)
library(tidyr)
library(limma)
library(tibble)
library(speckle)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/04_UBC_snRNAseq_vic_annotation/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

## Read in data

```{r}
vics <- readRDS(file = "../03_output/03_UBC_snRNAseq_atlas_rough_annotation/vics.rds")
```

# Initial clustering and QC

```{r}
# Use SelectIntegrationFeatures to select variable genes that are consistent
# across donors, reducing donor batch effects in the embedding
vics <- harmony_cluster(vics, dims = 1:50, use_integration_features = TRUE)
plot_cluster_umaps(vics)
```

```{r}
# Inspect QC metrics per cluster to identify low-quality or doublet clusters
plot_qc_by_cluster(vics, "RNA_snn_res.0.4")
```

```{r}
Idents(vics) <- "RNA_snn_res.0.4"
FAM(vics, "preqc_vics_0.4", output_path = output_path)
```

## QC filtering

RNA_snn_res.0.3 cluster 3 is very low quality; RNA_snn_res.0.3 cluster 8 is a doublet cluster; RNA_snn_res.0.4 cluster 4 has a very high percent of mitochondrial reads.

```{r}
dim(vics)
vics <- subset(vics, RNA_snn_res.0.4 != "4" & RNA_snn_res.0.4 != "5" & known_doublet == F & RNA_snn_res.0.4 != "10")
dim(vics)
```

# Re-cluster after filtering

```{r}
vics <- harmony_cluster(vics, dims = 1:50, use_integration_features = TRUE)
plot_cluster_umaps(vics)
```

```{r}
# Confirm QC metrics are clean after filtering
plot_qc_by_cluster(vics, "RNA_snn_res.0.3", show_doublets = FALSE)
```

# Annotation

```{r}
# Assign sequential cluster labels prefixed with "VICn" using resolution 0.3
vics$annotations_level2 <- as.numeric(vics$RNA_snn_res.0.3)
vics$annotations_level2[vics$RNA_snn_res.0.4 == "7"] <- 6

vics$annotations_level2 <- paste0("VICn", vics$annotations_level2)
vics$annotations_level2 <- factor(vics$annotations_level2, levels = sort(unique(vics$annotations_level2)))
vics$annotations_level2_readable <- mapvalues(vics$annotations_level2, from = c("VICn1", "VICn2", "VICn3", "VICn4", "VICn5", "VICn6", "VICn7"), to = c("Transitional", "Spongiosa", "FAP+ osteogenic", "Quiescent", "Contractile", "Neural crest-like", "IFN-stimulated"))
vics$annotations_level2_readable <- factor(vics$annotations_level2_readable, levels = rev(c("Transitional", "Quiescent", "Spongiosa", "FAP+ osteogenic", "Contractile", "Neural crest-like", "IFN-stimulated")))
DimPlot(vics, reduction = "umap", group.by = "annotations_level2", cols = "polychrome", raster = F) + coord_equal()
DimPlot(vics, reduction = "umap", group.by = "annotations_level2_readable", cols = "polychrome", raster = F) + coord_equal()
Idents(vics) <- "annotations_level2"
FAM(vics, "vics_annotations_level2", output_path = output_path)
Idents(vics) <- "annotations_level2_readable"
FAM(vics, "vics_annotations_level2_readable", output_path = output_path)
```

```{r}
### VIC Proportions Statistical Testing - Save Results for All Cell Types

# Step 1: Calculate proportions per sample for ALL cell types
proportions_df_all <- vics@meta.data %>%
  group_by(donor_id, clinical_classification, annotations_level2_readable) %>%
  summarise(n_cells = n(), .groups = "drop_last") %>%
  mutate(total_cells = sum(n_cells),
         proportion = n_cells / total_cells * 100) %>%  # Convert to percent
  ungroup()

# Step 2: Ensure all samples are present for all cell types (even if proportion = 0)
all_samples <- vics@meta.data %>%
  distinct(donor_id, clinical_classification)

all_celltypes <- unique(vics@meta.data$annotations_level2_readable)

# Create complete combination of samples and cell types
complete_combinations <- expand.grid(
  donor_id = all_samples$donor_id,
  annotations_level2_readable = all_celltypes,
  stringsAsFactors = FALSE
) %>%
  left_join(all_samples, by = "donor_id")

proportions_df_complete <- complete_combinations %>%
  left_join(proportions_df_all, by = c("donor_id", "clinical_classification", "annotations_level2_readable")) %>%
  mutate(proportion = replace_na(proportion, 0),
         n_cells = replace_na(n_cells, 0)) %>%
  dplyr::select(donor_id, clinical_classification, annotations_level2_readable, n_cells, total_cells, proportion)

# Step 3: Run statistical test
# Generate transformed proportions
props <- getTransformedProps(
  clusters = vics$annotations_level2_readable,
  sample = vics$donor_id,
  transform = "logit"
)

# Create sample info
sample_order <- colnames(props$TransformedProps)
sample_info <- vics@meta.data %>%
  distinct(donor_id, clinical_classification) %>%
  filter(donor_id %in% sample_order) %>%
  mutate(clinical_classification = factor(clinical_classification, levels = c("control", "mildmoderate", "severe")),
         dose = as.numeric(clinical_classification)) %>%
  arrange(factor(donor_id, levels = sample_order))

# Design matrix and fit
design <- model.matrix(~ dose, data = sample_info)
fit <- lmFit(props$TransformedProps, design)
fit <- eBayes(fit, robust = TRUE)

# Step 4: Extract p-values for ALL cell types
topTable_results_all <- topTable(fit, coef = "dose", number = Inf) %>%
  rownames_to_column("celltype") %>%
  arrange(adj.P.Val)

# Step 5: Save results to CSV
write.csv(topTable_results_all, 
          file = paste0(output_path, "VIC_proportions_statistics.csv"), 
          row.names = FALSE)

# Also save the proportions data
write.csv(proportions_df_complete, 
          file = paste0(output_path, "VIC_proportions_data.csv"), 
          row.names = FALSE)

# Print summary
cat("Statistical results saved for", nrow(topTable_results_all), "cell types\n")
cat("Significant cell types (adj.P.Val < 0.05):\n")
print(topTable_results_all %>% filter(adj.P.Val < 0.05) %>% dplyr::select(celltype, logFC, adj.P.Val))
```

```{r}
design <- model.matrix(~0 + clinical_classification, data = sample_info)
colnames(design) <- levels(sample_info$clinical_classification)

contrast_matrix <- makeContrasts(
  severe_specific = severe - (control + mildmoderate)/2,
  levels = design
)


fit <- lmFit(props$TransformedProps, design)
fit <- contrasts.fit(fit, contrast_matrix)
fit <- eBayes(fit, robust = TRUE)
severe_results <- topTable(fit, coef = "severe_specific", number = Inf) %>%
  rownames_to_column("celltype") %>%
  arrange(adj.P.Val)
write.csv(severe_results, 
          file = paste0(output_path, "VIC_proportions_severe_statistics.csv"), 
          row.names = FALSE)
```

# Pathway analysis

```{r}
universe <- rownames(vics)
DE_vics_annotations_level2 <- read.csv(paste0(output_path, "vics_annotations_level2_readable_degs.csv"), row.names = 1)

# Run GO biological process enrichment for each VIC subcluster and write to
# a multi-sheet workbook (one sheet per cluster)
wb <- createWorkbook()
for(i in unique(DE_vics_annotations_level2$cluster)){
  ego <- enrichGO(
    gene      = DE_vics_annotations_level2$gene[DE_vics_annotations_level2$cluster == i & DE_vics_annotations_level2$sig == "yes"],
    universe  = universe,
    OrgDb     = org.Hs.eg.db,
    keyType   = "SYMBOL",
    ont       = "BP",
    pvalueCutoff = 0.05,
    readable  = TRUE
  )
  addWorksheet(wb, i)
  writeData(wb, i, ego@result)
}
saveWorkbook(wb, file = paste0(output_path, "pathways_vics_annotations_level2_readable.xlsx"), overwrite = TRUE)
```

```{r}
emb <- Embeddings(vics, "umap")
emb[, "umap_1"] <- -emb[, "umap_1"]
emb[, "umap_2"] <- -emb[, "umap_2"]

vics[["umap"]]@cell.embeddings <- emb

DimPlot(vics, reduction = "umap", group.by = "annotations_level2_readable", cols = "polychrome", raster = F) + coord_equal()
```

# Export

```{r}
# Convert Seurat v5 Assay5 to legacy Assay for h5ad compatibility
vics_export <- vics
vics_export[["RNA3"]] <- as(object = vics_export[["RNA"]], Class = "Assay")
DefaultAssay(vics_export) <- "RNA3"
vics_export[["RNA"]] <- NULL
vics_export <- RenameAssays(object = vics_export, RNA3 = 'RNA')
anndataR::write_h5ad(vics_export, paste0(output_path, "vics.h5ad"), mode = "w")
```


```{r}
saveRDS(object = vics, file = paste0(output_path, "vics.rds"))
```

# MAGIC imputation

```{r}
# Write variable-gene count matrix for MAGIC imputation (run in Python)
magic_genes <- VariableFeatures(vics)
write10xCounts(
  path     = paste0(output_path, "vic_magic_input"),
  x        = vics@assays$RNA@layers$counts[which(rownames(vics) %in% magic_genes), ],
  barcodes = colnames(vics),
  gene.id  = rownames(vics)[which(rownames(vics) %in% magic_genes)],
  overwrite = T
)
```

```{r}
# Set up reticulate to use MAGIC mamba environment
library(reticulate)
use_condaenv("MAGIC", required = TRUE)
```

```{python}
# Run MAGIC on VIC object
import magic
import scprep
import numpy as np
import pandas as pd

vicnuc_data = scprep.io.load_10X("../03_output/04_UBC_snRNAseq_vic_annotation/vic_magic_input")
vicnuc_data = scprep.normalize.library_size_normalize(vicnuc_data)
vicnuc_data = scprep.transform.sqrt(vicnuc_data)
magic_operator = magic.MAGIC(random_state=42)
vicnuc_magic = magic_operator.fit_transform(vicnuc_data)
vicnuc_magic.to_csv("../03_output/04_UBC_snRNAseq_vic_annotation/vic_magic_output.csv")
```

```{r}
# Load MAGIC-imputed matrix and store as a separate assay for downstream use
vics_magic <- read.csv(paste0(output_path, "vic_magic_output.csv"), row.names = 1)
vics[["MAGIC"]] <- CreateAssayObject(data = t(vics_magic))
rm(vics_magic)
```

```{r}
saveRDS(object = vics, file = paste0(output_path, "vics_magic.rds"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path, 
                          "../01_data/output_checksums_04.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path, 
                          "../01_data/output_checksums_04.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```
