---
title: "07: UBC snRNA-seq Mononuclear Phagocyte Annotation"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to annotate the mononuclear phagocyte (MPS) cluster, which includes valve-resident macrophages, monocytes, and dendritic cells.

## Load packages

```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(harmony)
library(dplyr)
library(scuttle)
library(maditr)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/07_UBC_snRNAseq_myeloid_annotation/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

## Read in data

```{r}
myeloid <- readRDS(file = "../03_output/03_UBC_snRNAseq_atlas_rough_annotation/myeloid.rds")
```

# Initial clustering and QC

```{r}
# Use SelectIntegrationFeatures to reduce donor batch effects
myeloid <- harmony_cluster(myeloid, dims = 1:30, use_integration_features = TRUE)
plot_cluster_umaps(myeloid)
```

```{r}
# Inspect QC metrics per cluster to identify doublet/contaminating clusters
plot_qc_by_cluster(myeloid, "RNA_snn_res.0.2")
```

```{r}
Idents(myeloid) <- "RNA_snn_res.0.2"
FAM(myeloid, "preqc_myeloid_0.2", output_path = output_path)
```

## QC filtering

RNA_snn_res.0.2 cluster 3 is very low quality; RNA_snn_res.0.2 clusters 2 and 4 are mesenchymal doublet clusters.

```{r}
dim(myeloid)
myeloid <- subset(myeloid, RNA_snn_res.0.2 %in% c("2", "3", "4") | known_doublet == T, invert = T)
dim(myeloid)
```

# Re-cluster after filtering

```{r}
myeloid <- harmony_cluster(myeloid, dims = 1:30, use_integration_features = TRUE)
plot_cluster_umaps(myeloid)
```

```{r}
Idents(myeloid) <- "RNA_snn_res.0.2"
FAM(myeloid, "myeloid_0.2", output_path = output_path)
```

# Annotation

```{r}
# Assign sequential cluster labels prefixed with "Mn" using resolution 0.2
myeloid$annotations_level2 <- paste0("Mn", as.numeric(myeloid$RNA_snn_res.0.2))
myeloid$annotations_level2 <- factor(myeloid$annotations_level2, levels = sort(unique(myeloid$annotations_level2)))
DimPlot(myeloid, reduction = "umap", group.by = "annotations_level2", cols = "polychrome", raster = F) + coord_equal()
Idents(myeloid) <- "annotations_level2"
FAM(myeloid, "myeloid_annotations_level2", output_path = output_path)
```

# Save

```{r}
saveRDS(myeloid, paste0(output_path, "myeloid.rds"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path, 
                          "../01_data/output_checksums_07.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path, 
                          "../01_data/output_checksums_07.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```
