---
title: "08: UBC snRNA-seq Cycling Cells Annotation"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to annotate the cycling cell cluster. Cycling cells were identified at a higher resolution in the rough annotation (res 0.06) and span multiple cell lineages, annotated here by their parent cell type.

## Load packages

```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(harmony)
library(dplyr)
library(scuttle)
library(maditr)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/08_UBC_snRNAseq_cycling_annotation/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

## Read in data

```{r}
cycling <- readRDS(file = "../03_output/03_UBC_snRNAseq_atlas_rough_annotation/cycling.rds")
```

# Initial clustering and QC

```{r}
cycling <- harmony_cluster(cycling, dims = 1:30)
plot_cluster_umaps(cycling)
```

```{r}
# Inspect QC metrics per cluster to identify doublet/contaminating clusters
plot_qc_by_cluster(cycling, "RNA_snn_res.0.3")
```

```{r}
Idents(cycling) <- "RNA_snn_res.0.3"
FAM(cycling, "preqc_cycling_0.3", output_path = output_path)
```

## QC filtering

Remove scDblFinder-predicted doublets; no additional resolution-based filtering required.

```{r}
dim(cycling)
cycling <- subset(cycling, known_doublet == F)
dim(cycling)
```

# Re-cluster after filtering

```{r}
# Include high resolutions (1, 2, 5) to resolve rare cycling T cells from cycling VICs
high_res <- c(0.02, 0.04, 0.06, 0.08, 0.1, 0.2, 0.3, 0.4, 1, 2, 5)
cycling <- harmony_cluster(cycling, dims = 1:30, resolutions = high_res)
plot_cluster_umaps(cycling, resolutions = high_res)
```

```{r}
Idents(cycling) <- "RNA_snn_res.0.3"
FAM(cycling, "cycling_0.3", output_path = output_path)
```

```{r}
# Compare with Cell Ranger automated predictions to guide lineage assignment
DimPlot(cycling, group.by = "predicted_coarse_cell_type", cols = "polychrome") + coord_equal()
DimPlot(cycling, group.by = "predicted_fine_cell_type") + coord_equal()
table(cycling$predicted_coarse_cell_type)
table(cycling$predicted_fine_cell_type)
```

```{r}
# Confirm QC metrics are clean after filtering
plot_qc_by_cluster(cycling, "RNA_snn_res.0.3", show_doublets = FALSE)
```

# Annotation

```{r}
# Assign parent cell type lineage based on marker gene expression
cycling$annotations_level1 <- "proliferating"
DimPlot(cycling, reduction = "umap", group.by = "annotations_level1", cols = "polychrome", raster = F) + coord_equal()

cycling$annotations_level2_readable <- "proliferating"
cycling$annotations_level2_readable[cycling$RNA_snn_res.0.3 == "0"] <- "proliferating-immune"
cycling$annotations_level2_readable[cycling$RNA_snn_res.0.3 == "1"] <- "proliferating-mesenchymal"
DimPlot(cycling, reduction = "umap", group.by = "annotations_level2_readable", cols = "polychrome", raster = F) + coord_equal()

# Fine-grained sequential labels
cycling$annotations_level2 <- paste0("cyclingn", as.numeric(cycling$RNA_snn_res.0.3))
cycling$annotations_level2 <- factor(cycling$annotations_level2, levels = sort(unique(cycling$annotations_level2)))
DimPlot(cycling, reduction = "umap", group.by = "annotations_level2", cols = "polychrome", raster = F) + coord_equal()
Idents(cycling) <- "annotations_level2"
FAM(cycling, "cycling_annotations_level2", output_path = output_path)
```

# Save

```{r}
saveRDS(cycling, paste0(output_path, "cycling.rds"))
```


# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path, 
                          "../01_data/output_checksums_08.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path, 
                          "../01_data/output_checksums_08.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```