---
title: "10: UBC snRNA-seq Mast Cell Annotation"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to annotate the mast cell cluster.

## Load packages

```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(harmony)
library(dplyr)
library(scuttle)
library(maditr)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/10_UBC_snRNAseq_mast_annotation/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

## Read in data

```{r}
mast <- readRDS(file = "../03_output/03_UBC_snRNAseq_atlas_rough_annotation/mast.rds")
```

# Initial clustering and QC

```{r}
mast <- harmony_cluster(mast, dims = 1:30)
plot_cluster_umaps(mast)
```

```{r}
# Inspect QC metrics per cluster to identify doublet/contaminating clusters
plot_qc_by_cluster(mast, "RNA_snn_res.0.3")
```

```{r}
Idents(mast) <- "RNA_snn_res.0.3"
FAM(mast, "preqc_mast_0.3", output_path = output_path)
```

## QC filtering

RNA_snn_res.0.3 cluster 0 is a mesenchymal doublet cluster cluster 2 is a myeloid doublet cluster.

```{r}
dim(mast)
mast <- subset(mast, RNA_snn_res.0.3 == 0 | RNA_snn_res.0.3 == 2 | known_doublet == T, invert = T)
dim(mast)
```

# Re-cluster after filtering

```{r}
mast <- harmony_cluster(mast, dims = 1:30)
plot_cluster_umaps(mast)
```

```{r}
# Confirm QC metrics are clean after filtering
plot_qc_by_cluster(mast, "RNA_snn_res.0.2", show_doublets = FALSE)
```

# Annotation

```{r}
# Assign sequential cluster labels prefixed with "mastn" using resolution 0.2
mast$annotations_level2 <- paste0("mastn", as.numeric(mast$RNA_snn_res.0.2))
mast$annotations_level2 <- factor(mast$annotations_level2, levels = sort(unique(mast$annotations_level2)))
DimPlot(mast, reduction = "umap", group.by = "annotations_level2", cols = "polychrome", raster = F) + coord_equal()
```

# Save

```{r}
saveRDS(mast, paste0(output_path, "mast.rds"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path, 
                          "../01_data/output_checksums_10.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path, 
                          "../01_data/output_checksums_10.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```