---
title: "11: UBC snRNA-seq Annotation Combining"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to merge the individually annotated and QC-filtered cell type objects into a single atlas, reintegrate across donors, and compute a final set of dimension reductions and cluster annotations.

## Load packages

```{r}
library(Seurat)
library(dittoSeq)
library(scico)
library(ggplot2)
library(DropletUtils)
library(plyr)
library(harmony)
library(dplyr)
library(scuttle)
library(maditr)
library(tidyr)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/11_UBC_snRNAseq_annotation_combining/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

## Read in annotated objects

```{r}
# Load the QC-filtered, annotated object for each cell type
atlas_level1_initial_clusters <- c("vics", "myeloid", "tnkb_cells", "endothelial", "adipocytes", "mast", "cycling")
atlas_level1_initial_clusters_folders <- c("04_UBC_snRNAseq_vic_annotation", "07_UBC_snRNAseq_myeloid_annotation", "06_UBC_snRNAseq_tnkb_cells_annotation",
                                           "05_UBC_snRNAseq_endothelial_annotation", "09_UBC_snRNAseq_adipocytes_annotation",
                                           "10_UBC_snRNAseq_mast_annotation", "08_UBC_snRNAseq_cycling_annotation")
obj_list <- list()
for(i in 1:length(atlas_level1_initial_clusters)){
  obj_list[[length(obj_list) + 1]] <- readRDS(
    paste0("../03_output/", atlas_level1_initial_clusters_folders[i], "/", atlas_level1_initial_clusters[i], ".rds")
  )
}
atlas <- merge(obj_list[[1]], obj_list[-1])
```

# Reprocess combined atlas

```{r}
# Rejoin per-sample layers created on merge, then renormalise and reintegrate
# using SelectIntegrationFeatures for consistent variable gene selection across
# all cell types and donors
atlas <- JoinLayers(atlas)
atlas <- NormalizeData(atlas)
atlas <- harmony_cluster(atlas, dims = 1:50, use_integration_features = TRUE)
plot_cluster_umaps(atlas)
```

## Preview annotations carried over from subclustering notebooks

```{r}
DimPlot(atlas, reduction = "umap", group.by = "annotations_level1", cols = "polychrome", raster = F) + coord_equal()
DimPlot(atlas, reduction = "umap", group.by = "annotations_level2", cols = "polychrome", raster = F) + coord_equal()
```

```{r}
atlas$annotations_level1_readable <- mapvalues(atlas$annotations_level1, from = c("vics", "myeloid", "B_cell", "T_cell", "NK_cell", "endothelial", "adipocytes", "mast", "proliferating"), to = c("VICs", "Myeloid", "B cells", "T cells", "NK cells", "Endothelial cells", "Adipocytes", "Mast cells", "Proliferating"))
DimPlot(atlas, reduction = "umap", group.by = "annotations_level1_readable", cols = "polychrome", raster = F) + coord_equal()
```


```{r}
Idents(atlas) <- "annotations_level2"
FAM(atlas, "atlas_annotations_level2", output_path = output_path)
```

# Aortic stenosis gene signature enrichment

```{r}
# Derive an AS gene score from bulk RNA-seq DEGs and project onto single-cell data
bulk_DE_AS_v_noAS <- read.csv("../../01_bulk_seq/03_output/01_processing_bulkRNAseq/DE_AS_v_noAS.csv", row.names = 1)
up_genes <- subset(bulk_DE_AS_v_noAS, padj < 0.05 & log2FoldChange > 1)$hgnc_symbol
atlas <- AddModuleScore(atlas, features = list(up_genes = up_genes))
FeaturePlot(atlas, features = "Cluster1", max.cutoff = "q99") +
  scale_color_scico(palette = "vik", direction = 1, midpoint = 0,
                    limits = c(min(FetchData(atlas, "Cluster1")),
                               quantile(unlist(FetchData(atlas, "Cluster1")), 0.99))) +
  coord_equal()
```

```{r}
# Pairwise Wilcoxon test to compare AS gene score across cell types
pw <- pairwise.wilcox.test(atlas$Cluster1, atlas$annotations_level1, p.adjust.method = "BH")

pw_df <- as.data.frame(as.table(pw$p.value)) %>%
  filter(!is.na(Freq)) %>%
  rename(group1 = Var1, group2 = Var2, p_adj = Freq) %>%
  mutate(p_label = case_when(
    p_adj < 0.001 ~ "p<0.001",
    TRUE ~ paste0("p=", signif(p_adj, 3))
  ))

write.csv(pw_df, paste0(output_path, "Supplementary_Table_CellType_Signature_Pairwise.csv"), row.names = FALSE)
```

```{r}
Idents(atlas) <- "annotations_level1"
FAM(atlas, "atlas_annotations_level1", output_path = output_path)
```

# Save

```{r}
saveRDS(atlas, file = paste0(output_path, "atlas.rds"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path, 
                          "../01_data/output_checksums_11.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path, 
                          "../01_data/output_checksums_11.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```