---
title: "01: SALTIRE3 Xenium Preprocessing"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to load and preprocess SALTIRE3 Xenium spatial
transcriptomics data, perform cell-level QC filtering, integrate across donors with
Harmony, and transfer cell-type annotations from the matched scRNA-seq atlas using
SingleR.

## Load packages

```{r}
library(Seurat)
library(BPCells)
library(DropletUtils)
library(ggplot2)
library(dplyr)
library(plyr)
library(dittoSeq)
library(SingleCellExperiment)
library(SingleR)
library(scater)
library(tidyverse)
library(scico)
library(harmony)
library(future)
library(readxl)
library(plotly)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/01_SALTIRE3_xenium_preprocessing/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

# Load and merge Xenium data

## Read in sample metadata

```{r}
sample_info <- as.data.frame(read_excel("../01_data/sample_info.xlsx"))
obj_list <- list()
for(i in rownames(sample_info)){
  obj_list[[length(obj_list)+1]] <- LoadXenium(data.dir = paste0("../01_data/20251029__151959__MBSV20251027/", sample_info[i, "sample_id"]), fov = sample_info[i, "slide_id"])
  obj_list[[length(obj_list)]]$sample_id <- sample_info[i, "subsample_id"]
  obj_list[[length(obj_list)]]$slide_id <- sample_info[i, "slide_id"]
  obj_list[[length(obj_list)]]$donor_id <- substring(sample_info[i, "subsample_id"], 1, 10)
  # Load matched protein panel from the same cell feature matrix
  obj_list[[length(obj_list)]][['Protein']] <- CreateAssayObject(Read10X_h5(paste0("../01_data/20251029__151959__MBSV20251027/", sample_info[i, "sample_id"], "/cell_feature_matrix.h5"))$`Protein Expression`)
  obj_list[[length(obj_list)]] <- RenameCells(obj_list[[length(obj_list)]], add.cell.id = sample_info[i, "slide_id"])
}
```

## Merge samples

```{r}
options(future.globals.maxSize = 8 * 1024^3)
atlas <- merge(obj_list[[1]], c(obj_list[2:length(obj_list)]))
atlas <- JoinLayers(atlas)
```

# QC filtering

```{r}
VlnPlot(atlas, features = c("nFeature_Xenium", "nCount_Xenium"), ncol = 2, pt.size = 0)
dim(atlas)
# Remove cells with zero transcript counts (empty segmentation masks)
atlas <- subset(atlas, nCount_Xenium > 0)
VlnPlot(atlas, features = c("nFeature_Xenium", "nCount_Xenium"), ncol = 2, pt.size = 0)
dim(atlas)
```

```{r}
saveRDS(atlas, paste0(output_path, "Xenium_prefiltering.rds"))
```

```{r}
# Interactive histograms to inspect count/feature distributions before filtering
plot_ly(x = atlas$nCount_Xenium, type = "histogram")
plot_ly(x = atlas$nFeature_Xenium, type = "histogram")
plot_ly(x = atlas$nCount_Protein, type = "histogram")
plot_ly(x = atlas$nFeature_Protein, type = "histogram")
plot_ly(x = atlas$nCount_BlankCodeword, type = "histogram")
plot_ly(x = atlas$nCount_ControlProbe, type = "histogram")
```

```{r}
dim(atlas)
VlnPlot(atlas, features = c("nFeature_Xenium", "nCount_Xenium"), ncol = 2, pt.size = 0)
# Apply count and feature lower thresholds
atlas_lim <- subset(atlas, subset = nCount_Xenium < 10 | nFeature_Xenium < 5, invert = T)
VlnPlot(atlas_lim, features = c("nFeature_Xenium", "nCount_Xenium"), ncol = 2, pt.size = 0)
dim(atlas_lim)
```

# Normalise and identify variable features

```{r}
atlas_lim <- NormalizeData(atlas_lim)
atlas_lim <- FindVariableFeatures(atlas_lim)
topRNA <- head(VariableFeatures(atlas_lim), 100)

# Plot variable features with and without labels
plot1RNA <- VariableFeaturePlot(atlas_lim)
plot2RNA <- LabelPoints(plot = plot1RNA, points = topRNA, repel = TRUE)
plot2RNA
```

```{r}
atlas_lim <- NormalizeData(atlas_lim, assay = "Protein")
atlas_lim <- FindVariableFeatures(atlas_lim, assay = "Protein")
topProtein <- head(VariableFeatures(atlas_lim, assay = "Protein"), 100)

# Plot variable features with and without labels
plot1Protein <- VariableFeaturePlot(atlas_lim, assay = "Protein")
plot2Protein <- LabelPoints(plot = plot1Protein, points = topProtein, repel = TRUE)
plot2Protein
```

# Normalise, integrate, and dimensionally reduce counts matrix

```{r}
# Note: harmony_cluster() uses default assay/reduction names suitable for scRNA-seq;
# Xenium data requires Xenium-specific assay names and custom reduction keys,
# so the integration pipeline is run inline here.
atlas_lim <- atlas_lim %>%
  ScaleData(assay = "Xenium") %>%
  RunPCA(assay = "Xenium", reduction.name = "pca_rna") %>%
  RunHarmony(group.by.vars = "donor_id", reduction.use = "pca_rna") %>%
  RunUMAP(reduction = "harmony", dims = 1:30, reduction.name = "umap_rna") %>%
  FindNeighbors(reduction = "harmony", dims = 1:30, graph.name = "rna_snn") %>%
  FindClusters(resolution = c(0.02, 0.04, 0.06, 0.08, 0.1, 0.2), graph.name = "rna_snn")
DimPlot(atlas_lim, cols = dittoColors(), group.by = "donor_id", raster=FALSE)+coord_equal()
DimPlot(atlas_lim, cols = dittoColors(), group.by = "sample_id", raster=FALSE)+coord_equal()
DimPlot(atlas_lim, group.by = "rna_snn_res.0.02", cols = dittoColors(), raster=FALSE, reduction = "umap_rna")+coord_equal()
DimPlot(atlas_lim, group.by = "rna_snn_res.0.04", cols = dittoColors(), raster=FALSE, reduction = "umap_rna")+coord_equal()
DimPlot(atlas_lim, group.by = "rna_snn_res.0.06", cols = dittoColors(), raster=FALSE, reduction = "umap_rna")+coord_equal()
DimPlot(atlas_lim, group.by = "rna_snn_res.0.08", cols = dittoColors(), raster=FALSE, reduction = "umap_rna")+coord_equal()
DimPlot(atlas_lim, group.by = "rna_snn_res.0.1", cols = dittoColors(), raster=FALSE, reduction = "umap_rna")+coord_equal()
DimPlot(atlas_lim, group.by = "rna_snn_res.0.2", cols = dittoColors(), raster=FALSE, reduction = "umap_rna")+coord_equal()
```

# Annotate cells using SingleR transfer from scRNA-seq atlas

```{r}
single_cell_atlas <- readRDS("../../03_SALTIRE3_scRNA_seq/03_output/10_S3_scRNAseq_annotations_combining/merged_objects.rds")
```

```{r}
unique(single_cell_atlas$annotations_level3_readable)
single_cell_atlas_sce <- as.SingleCellExperiment(single_cell_atlas)
atlas_sce <- as.SingleCellExperiment(atlas_lim)
```

```{r, eval=FALSE}
pred <- SingleR(test=atlas_sce, ref=single_cell_atlas_sce,  labels=single_cell_atlas_sce$annotations_level3_readable, de.method="wilcox", BPPARAM = BiocParallel::MulticoreParam(),
  fine.tune = FALSE,
  prune = FALSE)
write.csv(pred, paste0(output_path, "pred.csv"))
```

```{r}
pred <- read.csv(paste0(output_path, "pred.csv"))
atlas_lim[["labels"]] <- pred$labels
```

```{r}
DimPlot(atlas_lim, group.by = "labels", raster = F, cols = dittoColors(), label = T)+coord_equal()
```

```{r}
# Map SingleR labels to annotation hierarchy using reference table
annotation_reference <- read.csv("../01_data/annotation_reference.csv")
atlas_lim$annotations_level3 <- atlas_lim$labels
atlas_lim$annotations_level2 <- mapvalues(atlas_lim$annotations_level3, from = annotation_reference$annotations_level3, to = annotation_reference$annotations_level2)
atlas_lim$annotations_level1 <- mapvalues(atlas_lim$annotations_level3, from = annotation_reference$annotations_level3, to = annotation_reference$annotations_level1)
DimPlot(atlas_lim, group.by = "annotations_level3", raster = F, cols = dittoColors(), label = T)+coord_equal()
DimPlot(atlas_lim, group.by = "annotations_level2", raster = F, cols = dittoColors(), label = T)+coord_equal()
DimPlot(atlas_lim, group.by = "annotations_level1", raster = F, cols = dittoColors(), label = T)+coord_equal()
```
```{r}
ImageDimPlot(atlas_lim, fov = "BI5XO_A1", group.by = "annotations_level2", cols = "polychrome")+coord_equal()+theme_classic()
```

```{r}
df_annotations <- data.frame(row.names = colnames(atlas_lim), annotations_level1 = atlas_lim$annotations_level1, annotations_level2 = atlas_lim$annotations_level2, annotations_level3 = atlas_lim$annotations_level3)
write.csv(df_annotations, file = paste0(output_path, "xenium_annotations.csv"))
```

# Save

```{r}
saveRDS(atlas_lim, paste0(output_path, "atlas_lim.rds"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path,
                          "../01_data/output_checksums_01.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path,
                          "../01_data/output_checksums_01.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```
