---
title: "02: SALTIRE3 Xenium histology"
author: "Stefan Veizades"
license: "MIT"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center',
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to assess the histology of sections adjacent
to the Xenium capture areas, align histological annotations to Xenium coordinates,
and compute per-cell distances to annotated tissue structures.

## Load packages

```{r}
library(Seurat)
library(ggplot2)
library(dplyr)
library(readxl)
library(pheatmap)
library(sf)
library(scico)
library(dittoSeq)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/02_SALTIRE3_xenium_histology/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")

# Helper functions for affine-transforming QuPath annotation geometries
# into Xenium pixel coordinates
affine_xy <- function(xy, M) {
  # xy: n x 2 matrix; M: 3x3 affine transformation matrix
  out <- cbind(xy, 1) %*% t(M)
  out[, 1:2, drop = FALSE]
}

# Apply affine transformation to an sf POLYGON/MULTIPOLYGON geometry,
# optionally rescaling by pixel_size (micrometres per pixel)
affine_geom <- function(g, M, pixel_size = NULL) {
  g <- sf::st_cast(g, "MULTIPOLYGON")

  xy <- sf::st_coordinates(g)
  xy2 <- affine_xy(xy[, c("X","Y"), drop = FALSE], M)

  if (!is.null(pixel_size)) {
    xy2 <- xy2 * pixel_size
  }

  # Rebuild MULTIPOLYGON structure using L1 (polygon id) and L2 (ring id)
  df <- data.frame(
    X = xy2[,1], Y = xy2[,2],
    L1 = xy[, "L1"], L2 = xy[, "L2"]
  )

  polys <- lapply(split(df, df$L1), function(p1) {
    rings <- lapply(split(p1, p1$L2), function(r) {
      as.matrix(r[, c("X","Y"), drop = FALSE])
    })
    rings
  })

  sf::st_multipolygon(polys)
}
```

## Read in data

```{r}
atlas_lim <- readRDS("../03_output/01_SALTIRE3_xenium_preprocessing/atlas_lim.rds")
```

# Add scNiche assignments

```{r}
niches <- read.csv("../01_data/scNiche_cell_assignments.csv", row.names = 1)
```

```{r}
niches$scNiche <- plyr::mapvalues(niches$scNiche, from = paste0("Niche", 0:3), to = c("Cap", "Spongiosa", "Peri-nodular", "Immune"))
atlas_lim$niche <- niches$scNiche
DimPlot(atlas_lim, group.by = "niche", raster = F, cols = spatial_niche_cols)+coord_equal()
```

# Align histological annotations and compute distances to nodule edge

```{r}
annotations <- rbind(read_sf("../01_data/E251106C_BI5XO-A10_PH2N_Scan1_all_annotations.geojson"), read_sf("../01_data/E251106C_BXNFB-A10_PH2N_Scan1.geojson"))
annotations <- annotations[!is.na(annotations$name),]
annotations$name <- gsub(x = annotations$name, pattern = "B15XO", replacement = "BI5XO")
```

```{r}
pixel_size <- 0.2125
atlas_lim$dist_to_nodule_edge <- 999
for(i in unique(atlas_lim$slide_id)){
  # Read affine transformation matrix from QuPath-Xenium alignment
  M <- matrix(scan(paste0("../01_data/Post_Xenium_IF_alignment_files_", i, "/matrix.csv"), sep=","), nrow=3, byrow=TRUE)

  # Extract nodular tissue annotation polygon for this slide
  sf_obj <- annotations[annotations$name == paste0(i, " Nodular Tissue"),]
  sf::st_crs(sf_obj) <- sf::st_crs(NA)

  # Transform nodular tissue polygon to Xenium coordinate space
  sf_obj_xenium <- sf_obj
  sf::st_geometry(sf_obj_xenium) <- sf::st_sfc(
    lapply(sf::st_geometry(sf_obj), affine_geom, M = M, pixel_size = pixel_size),
    crs = sf::st_crs(NA)
  )

  # Get cell coordinates for this slide
  coords <- GetTissueCoordinates(atlas_lim, image = i)
  cells_sf <- st_as_sf(
    coords,
    coords = c("x", "y"),
    crs = NA
  )

  # Calculate per-cell distance to nodule boundary
  d <- st_distance(cells_sf, sf_obj_xenium)

  atlas_lim$dist_to_nodule_edge[atlas_lim$slide_id == i] <- as.numeric(d)
}
```

# Save

```{r}
saveRDS(atlas_lim, paste0(output_path, "atlas_lim.rds"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path,
                          "../01_data/output_checksums_02.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path,
                          "../01_data/output_checksums_02.csv")
```

```{r}
sessioninfo::session_info(info = "packages")
```
