---
title: "01: Bulk RNA-seq Data Analysis"
author: "Stefan Veizades"
license: "MIT"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to perform differential gene expression and pathway analysis of bulk RNA-seq data generated using Lexogen's QuantSeq FWD. Samples are compared across clinical classifications (control, mild/moderate, severe aortic stenosis) using DESeq2.

## Load packages

```{r}
library(dplyr)
library(readxl)
library(DESeq2)
library(ggplot2)
library(GSVA)
library(org.Hs.eg.db)
library(clusterProfiler)
library(biomaRt)
library(scico)
library(digest)
```

## Parameters

```{r}
# Paths
output_path <- "../03_output/01_processing_bulkRNAseq/"
if(!dir.exists(output_path)) dir.create(output_path)

# Functions
source("../../utils.R")
```

# Format sample and sequencing data for DESeq2

```{r}
# Sample and donor metadata
sample_info <- as.data.frame(read_excel("../../00_study_design/quantseq_info.xlsx"))
donor_info <- as.data.frame(read_excel("../../00_study_design/patient_information.xlsx"))
rownames(donor_info) <- donor_info$donor_id
sample_donor_info <- left_join(x = sample_info, y = donor_info, by = 'donor_id')
sample_list <- paste0(sample_donor_info$sample_id[sample_donor_info$bulk_seq == "yes"], "/", sample_donor_info$sample_id[sample_donor_info$bulk_seq == "yes"], "_gene_featureCounts_output.txt")
write.table(sample_list, file = paste0(output_path, "featureCounts_file_list.txt"), quote = F, row.names = F, col.names = F)
```

```{bash}
# Retrieve sequencing data from NAS
source ~/.bashrc
rsync -av -e "ssh -p 22 -i ~/.ssh/id_coconut -o IdentitiesOnly=yes" \
  --no-R --files-from=../03_output/01_processing_bulkRNAseq/featureCounts_file_list.txt \
  coconut:"$COCONUT_CUREAS_PATH"Results/01-bulk-seq_Processing/04-featureCounts \
  ../01_data/featureCounts_output
```

```{r, eval=FALSE}
# Only run once, to generate initial checksum files
# Read-in loop with checksum generation
counts <- read.delim(
  paste0("../01_data/featureCounts_output/", 
         sample_donor_info[1, "sample_id"], 
         "_gene_featureCounts_output.txt"), 
  comment.char = "#"
)
aggregated_counts <- data.frame(row.names = counts$Geneid)
checksums <- list()

for (i in rownames(sample_donor_info[sample_donor_info$bulk_seq == "yes", ])) {
  sid <- sample_donor_info[i, "sample_id"]
  counts <- read.delim(
    paste0("../01_data/featureCounts_output/", sid, 
           "_gene_featureCounts_output.txt"),
    comment.char = "#"
  )
  
  checksums[[sid]] <- checksum_featurecounts(counts)
  message(paste0(sid, ": ", checksums[[sid]]))
  
  rownames(counts) <- counts$Geneid
  aggregated_counts[rownames(counts), sample_donor_info[i, "donor_id"]] <- 
    counts[, 7:ncol(counts)]
}

rownames(aggregated_counts) <- gsub("\\..*", "", rownames(aggregated_counts))

# Save reference checksums
checksum_df <- data.frame(md5 = unlist(checksums), row.names = names(checksums))
write.csv(checksum_df, "../01_data/featureCounts_checksums.csv")
message("Reference checksums saved to ../01_data/featureCounts_checksums.csv")
```


```{r, message=TRUE}
# Read in reference checksums
reference_checksums <- read.csv("../01_data/featureCounts_checksums.csv", 
                                row.names = 1)

# Read-in loop
# Get list of genes from first sample to serve as row names for data frame of sample counts
counts <- read.delim(
  paste0("../01_data/featureCounts_output/", 
         sample_donor_info[1, "sample_id"], 
         "_gene_featureCounts_output.txt"), 
  comment.char = "#"
)
aggregated_counts <- data.frame(row.names = counts$Geneid)
checksums <- list()

# Read in count data for each sample and add to aggregated matrix
for (i in rownames(sample_donor_info[sample_donor_info$bulk_seq == "yes", ])) {
  sid <- sample_donor_info[i, "sample_id"]
  counts <- read.delim(
    paste0("../01_data/featureCounts_output/", sid, 
           "_gene_featureCounts_output.txt"),
    comment.char = "#"
  )
  
  # Validate before using
  checksums[[sid]] <- validate_featurecounts(counts, sid, reference_checksums)
  
  rownames(counts) <- counts$Geneid
  aggregated_counts[rownames(counts), sample_donor_info[i, "donor_id"]] <- 
    counts[, 7:ncol(counts)]
}

rownames(aggregated_counts) <- gsub("\\..*", "", rownames(aggregated_counts))
```

```{r}
# Format sample meta data for DESeq2
coldata <- data.frame(
  row.names = colnames(aggregated_counts),
  clinical_classification = donor_info[colnames(aggregated_counts), "clinical_classification"],
  aortic_stenosis_diagnosis = donor_info[colnames(aggregated_counts), "aortic_stenosis_diagnosis"])
```

# DESeq2

## Import data into DESeq2

```{r}
dds <- DESeqDataSetFromMatrix(countData = aggregated_counts,
                              colData = coldata,
                              design = ~ clinical_classification)
dds <- dds[rowSums(counts(dds)) > 10, ] # Remove genes with less than 10 counts across samples

# Create second DESeq2 object to run aortic stenosis vs control analysis
dds_as <- DESeqDataSetFromMatrix(countData = aggregated_counts,
                              colData = coldata,
                              design = ~ aortic_stenosis_diagnosis)
dds_as <- dds_as[rowSums(counts(dds_as)) > 10, ] # Remove genes with less than 10 counts across samples
```

## Run principal component analysis

```{r}
# Normalise counts
vsd <- vst(dds, blind=FALSE)

# Generate principal component analysis plot
pca_plot <- plotPCA(vsd, intgroup=c("clinical_classification"))
pca_plot
plotPCA(vsd, intgroup=c("aortic_stenosis_diagnosis"))
pca_plot_data <- pca_plot$data
pca_plot_data$group <- NULL
pca_plot_data$aortic_stenosis_diagnosis <- coldata[rownames(pca_plot_data), "aortic_stenosis_diagnosis"]

# Save VSD
saveRDS(vsd, file = paste0(output_path, "bulk_vsd.rds"))
```

# Differential gene expression analysis

## Differential gene testing

```{r}
# Prepare DESeq
dds <- DESeq(dds)
dds_as <- DESeq(dds_as)

# Run tests
sev_v_mm <- results(dds, contrast = c("clinical_classification", "severe", "mildmoderate"))
sev_v_ctrl <- results(dds, contrast = c("clinical_classification", "severe", "control"))
mm_v_ctrl <- results(dds, contrast = c("clinical_classification", "mildmoderate", "control"))
AS_v_noAS <- results(dds_as, contrast = c("aortic_stenosis_diagnosis", "AS", "noAS"))

# Convert Ensembl gene to gene IDs
genes <- rownames(dds)
annot_file <- paste0(output_path, "ensembl_annotations.csv")
if (file.exists(annot_file)) {
  annot <- read.csv(annot_file)
} else {
  mart <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl", host = "www.ensembl.org") 
  annot <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"), filters = "ensembl_gene_id", values = genes,mart = mart)
  write.csv(annot, annot_file, row.names = FALSE)
}
annot$hgnc_symbol[annot$hgnc_symbol == ""] <- annot$ensembl_gene_id

# Add column with gene IDs
sev_v_mm$ensembl_gene_id <- rownames(sev_v_mm)
sev_v_mm <- merge(as.data.frame(sev_v_mm), annot, by = "ensembl_gene_id", all.x = TRUE)
sev_v_ctrl$ensembl_gene_id <- rownames(sev_v_ctrl)
sev_v_ctrl <- merge(as.data.frame(sev_v_ctrl), annot, by = "ensembl_gene_id", all.x = TRUE)
mm_v_ctrl$ensembl_gene_id <- rownames(mm_v_ctrl)
mm_v_ctrl <- merge(as.data.frame(mm_v_ctrl), annot, by = "ensembl_gene_id", all.x = TRUE)
AS_v_noAS$ensembl_gene_id <- rownames(AS_v_noAS)
AS_v_noAS <- merge(as.data.frame(AS_v_noAS), annot, by = "ensembl_gene_id", all.x = TRUE)
sev_v_mm$hgnc_symbol[is.na(sev_v_mm$hgnc_symbol)] <- sev_v_mm$ensembl_gene_id
sev_v_ctrl$hgnc_symbol[is.na(sev_v_ctrl$hgnc_symbol)] <- sev_v_ctrl$ensembl_gene_id
mm_v_ctrl$hgnc_symbol[is.na(mm_v_ctrl$hgnc_symbol)] <- mm_v_ctrl$ensembl_gene_id
AS_v_noAS$hgnc_symbol[is.na(AS_v_noAS$hgnc_symbol)] <- AS_v_noAS$ensembl_gene_id

# Write DE results
write.csv(sev_v_mm, file = paste0(output_path, "DE_sev_v_mm.csv"))
write.csv(sev_v_ctrl, file = paste0(output_path, "DE_sev_v_ctrl.csv"))
write.csv(mm_v_ctrl, file = paste0(output_path, "DE_mm_v_ctrl.csv"))
write.csv(AS_v_noAS, file = paste0(output_path, "DE_AS_v_noAS.csv"))

# Save DESeq2 objects
saveRDS(dds, file = paste0(output_path, "DESeq2_condition_obj.rds"))
saveRDS(dds_as, file = paste0(output_path, "DESeq2_as_obj.rds"))
saveRDS(dds_as, file = paste0(output_path, "DESeq2_as_obj.rds"))
```

## Find aortic stenosis gene score

```{r}
# Subset AS v control DE for genes significantly up in AS and log 2 FC > 1
up_genes <- subset(AS_v_noAS, padj < 0.05 & log2FoldChange > 1)$ensembl_gene_id

# Get GSVA enrichment scores per sample
gbmPar <- gsvaParam(assay(vsd), list(up_genes=up_genes), maxDiff=FALSE) 
gbm_es <- gsva(gbmPar)
```

```{r}
# Get principal component analysis plot with aortic stenosis gene score
pca_plot_data$gene_score <- as.numeric(gbm_es)
ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = gene_score)) +
  geom_point(size = 3) +
  scale_color_scico(palette = "vik", midpoint = 0, direction = 1) +
  theme_minimal() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
write.csv(pca_plot_data, file = paste0(output_path, "pca_data.csv"))
```

## Pathway analysis

```{r}
# Set background gene list to all genes in bulk seq data
universe <- rownames(dds)

# Pathways upregulated in severe
ego_sev_v_ctrl <- enrichGO(gene = sev_v_ctrl[sev_v_ctrl$padj < 0.05 & !is.na(sev_v_ctrl$padj) & sev_v_ctrl$log2FoldChange > 1, "ensembl_gene_id"], universe = universe, OrgDb = org.Hs.eg.db, keyType = "ENSEMBL", ont = "BP", pvalueCutoff = 0.05, readable = TRUE)
write.csv(ego_sev_v_ctrl@result, file = paste0(output_path, "severe_pathways.csv"))

# Pathways upregulated in control
ego_ctrl_v_sev_or_ctrl_v_mm <- enrichGO(gene = unique(sev_v_ctrl[sev_v_ctrl$padj < 0.05 & !is.na(sev_v_ctrl$padj) & sev_v_ctrl$log2FoldChange < -1, "ensembl_gene_id"], mm_v_ctrl[mm_v_ctrl$padj < 0.05 & !is.na(mm_v_ctrl$padj) & mm_v_ctrl$log2FoldChange < -1, "ensembl_gene_id"]), universe = universe, OrgDb = org.Hs.eg.db, keyType = "ENSEMBL", ont = "BP", pvalueCutoff = 0.05, readable = TRUE)
write.csv(ego_ctrl_v_sev_or_ctrl_v_mm@result, file = paste0(output_path, "control_pathways.csv"))

# Pathways upregulated in mild/moderate
ego_mm_v_ctrl <- enrichGO(gene = mm_v_ctrl[mm_v_ctrl$padj < 0.05 & !is.na(mm_v_ctrl$padj) & mm_v_ctrl$log2FoldChange > 1, "ensembl_gene_id"], universe = universe, OrgDb = org.Hs.eg.db, keyType = "ENSEMBL", ont = "BP", pvalueCutoff = 0.05, readable = TRUE)
write.csv(ego_mm_v_ctrl@result, file = paste0(output_path, "mildmoderate_pathways.csv"))
```

# --- Run once to generate reference checksums ---

```{r, eval=FALSE}
generate_output_checksums(output_path, 
                          "../01_data/output_checksums_01.csv")
```

# --- Run on every subsequent execution ---

```{r, message=TRUE}
validate_output_checksums(output_path, 
                          "../01_data/output_checksums_01.csv")
```

```{r}
package_versions <- as.data.frame(sessioninfo::package_info())
package_versions[, intersect(c("package", "loadedversion", "date", "source"), names(package_versions))]
```
